---
title: "Differential gene expression of pancreatic cancer for the dataset GSE183795"
author: "Iswarya L U"
date: "5-11-2025"
output: html_document
---
##Necessary packages
```{r}
library(limma)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
library(pheatmap)
library(scales)
library(ggrepel)

```

**rm(list = ls())** clears the R environment to start with a clean workspace.
**limma** is used for differential gene expression analysis.
**tidyverse** is used for data manipulation.
**clusterProfiler** is used for pathway and enrichment analysis.
**org.Hs.eg.db** provides human gene annotation.
**enrichplot and pheatmap** are used for enrichment and heatmap visualization.
**ggplot2** is used for plotting

## 1. Dataset and Study Design

This analysis uses the publicly available GEO dataset GSE183795, 
containing gene-expression profiles from pancreatic ductal 
adenocarcinoma (PDAC) tumor and adjacent non-tumor tissues.

The dataset contains 139 tumor samples, 102 adjacent non-tumor 
samples, and 3 normal pancreas samples from donors.

For this analysis, only tumor and adjacent non-tumor samples were 
included. The three normal donor samples were excluded because the 
objective of this analysis is to identify transcriptional differences 
between PDAC tumor and adjacent non-tumor tissues.

The final analysis therefore includes 241 samples:
139 tumor and 102 adjacent non-tumor samples.

##2. Load Processed Data
```{r}
expr <- read.csv("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/normalized_expression_matrix.csv",row.names = 1, check.names = FALSE)

meta <- read.csv("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/meta_clean.csv",row.names = 1)

```

**read.csv()** imports the preprocessed expression matrix and metadata.
**row.names = 1** uses the first column as row names.
**check.names = FALSE** preserves the original sample names.## Including Plots

##3. Select Tumor and Adjacent Non-Tumor Samples

```{r}
colnames(expr) <- gsub("_.*", "", colnames(expr))

meta_dge <- meta %>%
  filter(tissue %in% c("Tumor", "adjacent_non_tumor"))

expr_dge <- expr[, rownames(meta_dge)]

```

**gsub()** standardizes the expression matrix sample names.
**filter()** retains only tumor and adjacent non-tumor samples.
The expression matrix is then subsetted to match the selected metadata samples.

The three normal donor samples are excluded because this analysis focuses on tumor vs adjacent non-tumor tissue.

##4. Define Experimental Groups
```{r}
group <- factor(
  meta_dge$tissue,
  levels = c("adjacent_non_tumor", "Tumor")
)

design <- model.matrix(~0 + group)
colnames(design) <- levels(group)
```
**factor()**defines the two experimental groups.
**model.matrix()** creates the design matrix required for the limma analysis.

Adjacent non-tumor tissue is used as the reference group.

##5.Differential Gene Expression Analysis

```{r}
fit <- lmFit(expr_dge, design)

contrast.matrix <- makeContrasts(
  Tumor - adjacent_non_tumor,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
```
**lmFit()** fits a linear model for each gene.
**makeContrasts()** defines the Tumor vs Adjacent Non-Tumor comparison.
**contrasts.fit()** applies this comparison to the fitted model.
**eBayes()** performs empirical Bayes moderation to improve variance estimation.

##6.Extract DEG Results

```{r}
deg_results <- topTable(
  fit2,
  number = Inf,
  adjust.method = "BH"
) %>%
  rownames_to_column("Gene")
```
**topTable()** extracts differential expression statistics for all genes.
number = Inf retains all genes.
BH applies Benjamini-Hochberg multiple-testing correction.
**rownames_to_column()** converts gene IDs into a Gene column.

##7. Classify DEGs

```{r}
deg_results <- deg_results %>%
  mutate(
    status = case_when(
      adj.P.Val < 0.05 & logFC > 1 ~ "Upregulated",
      adj.P.Val < 0.05 & logFC < -1 ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
)
write.csv(deg_results,"C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/csv_of_differentially_expressed_genes.csv", row.names = FALSE)
```

Genes are classified using:

**Adjusted p-value < 0.05**
**log2FC > 1 → Upregulated**
**log2FC < -1 → Downregulated**
**Remaining genes → Not Significant**

##8.Volcano Plot

```{r}
volcano_plot <- ggplot(deg_results, 
                       aes(x = logFC, 
                           y = -log10(adj.P.Val), 
                           color = status)) +
  geom_point(alpha = 0.7, size = 1.8) +
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  theme_bw(base_size = 12) +
  labs(title = "Volcano Plot: Tumor vs Adjacent Non-Tumor",
       x = "Log2 Fold Change",
       y = "-Log10 Adjusted P-value")

volcano_plot

ggsave(
  filename = "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/deg/DEG_Volcano_Plot.png",
  plot = volcano_plot,
  width = 12,
  height = 9,
  units = "in",
  dpi = 300
)

```

**ggplot()** initializes the plot.
**aes()** maps log2 fold change to the x-axis and adjusted p-value to the y-axis.
**geom_point()** displays individual genes.
**geom_vline() and geom_hline()** show the DEG thresholds.

##9.Top 30 DEG heatmap

```{r}
top30 <- deg_results %>%
  arrange(adj.P.Val) %>%
  filter(status != "Not Significant") %>%
  head(30)

top30_genes <- top30$Gene
expr_top30 <- expr_dge[top30_genes, ]
expr_scaled <- t(scale(t(expr_top30)))

annotation_col <- data.frame(
  Tissue = meta_dge$tissue
)

rownames(annotation_col) <- rownames(meta_dge)

ann_colors <- list(
  Tissue = c(
    Tumor = "firebrick3",
    adjacent_non_tumor = "steelblue3"
  )
)

pheatmap(
  expr_scaled,
  annotation_col = annotation_col,
  show_colnames = FALSE,
  fontsize_row = 8,
  clustering_method = "complete"
)

pheatmap(
  expr_scaled,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = FALSE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
  fontsize_row = 8,
  main = "Top 30 DEGs: Tumor (Red) vs Adjacent Non-Tumor (Blue)"
)

# Record the displayed heatmap
heatmap_recorded <- recordPlot()
# ============================================================
# SAVE HEATMAP
# ============================================================

figure_dir <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/deg"

if (!dir.exists(figure_dir)) {
  dir.create(figure_dir, recursive = TRUE)
}

png(
  filename = file.path(figure_dir, "DEG_Heatmap.png"),
  width = 4200,
  height = 3600,
  res = 300
)

replayPlot(heatmap_recorded)

dev.off()
```
