title: "Weighted Gene Co-expression Network Analysis of Pancreatic Cancer for the Dataset GSE183795"
author: "Iswarya L U"
date: "13-09-2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

# Weighted Gene Co-expression Network Analysis

## Overview

Weighted Gene Co-expression Network Analysis (WGCNA) was performed to identify
gene co-expression modules associated with pancreatic cancer status.

The analysis was performed using the normalized expression matrix from
GSE183795.

The original dataset contained approximately:

- 19,974 genes
- 244 samples
- 139 Tumor samples
- 102 adjacent non-tumor samples
- 3 Normal samples

Variance filtering was performed using all available samples so that the Normal
samples contributed to the identification of highly variable genes.

The Normal samples were subsequently excluded from WGCNA network construction.
Therefore, the final WGCNA network was constructed using Tumor and adjacent
non-tumor samples only.

Genes above the 25th percentile of gene variance were retained for downstream
network analysis.

The major steps of this analysis were:

1. Loading required R packages
2. Loading normalized expression and metadata
3. Checking tissue groups
4. Aligning expression data and metadata
5. Variance filtering
6. Selecting Tumor and adjacent non-tumor samples
7. Preparing the WGCNA expression matrix
8. Quality control
9. Sample clustering
10. Principal component analysis
11. Soft-threshold power selection
12. Network construction
13. Module identification
14. Module-trait relationship analysis
15. Module membership and gene significance analysis
16. Identification of significant modules
17. Identification of hub genes
18. Saving WGCNA results

---

# 1. Necessary Packages

This section loads all R packages required for the WGCNA analysis,
data manipulation, visualization, and heatmap generation.

```{r setup, message=FALSE, warning=FALSE,fig.width=12 , fig.height=16,  dpi=100, out.width="1200px",out.height="1600px"}

# Clear the R environment
rm(list = ls())

# Close previously opened graphics devices
graphics.off()

# Set seed for reproducibility
set.seed(1234)

# Load required packages
library(WGCNA)
library(matrixStats)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(gridExtra)

# Prevent automatic conversion of strings to factors
options(stringsAsFactors = FALSE)

```

##Enable WGCNA Multithreading

WGCNA can use multiple CPU cores for computationally intensive operations.
This can considerably reduce the execution time for large expression matrices.


```{r}

# Allow WGCNA to use multiple CPU cores
enableWGCNAThreads()

```

# 2. Load Processed Expression Data and Metadata

The normalized expression matrix and corresponding sample metadata are loaded
from CSV files.

The expression matrix is expected to contain genes as rows and samples as
columns.

The metadata contains sample information, including the tissue classification.


```{r}

# Define input file paths

expr_path <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/normalized_expression_matrix.csv"

meta_path <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/meta_clean.csv"


# Load normalized expression matrix

expr <- read.csv(
  expr_path,
  row.names = 1,
  check.names = FALSE
)


# Load sample metadata

meta <- read.csv(
  meta_path,
  row.names = 1,
  check.names = FALSE
)


# Display dimensions

cat("Expression matrix dimensions:\n")
print(dim(expr))

cat("\nMetadata dimensions:\n")
print(dim(meta))


# Display first few genes

cat("\nFirst few expression genes:\n")
print(head(rownames(expr)))


# Display first few metadata rows

cat("\nFirst few metadata rows:\n")
print(head(meta))


```

# 3. Check Tissue Groups

Before performing WGCNA, the distribution of samples among the different tissue
groups is examined.

This confirms that the metadata contains the expected sample classifications.


```{r}

cat("Tissue groups:\n")

print(
  table(meta$tissue)
)


```

The expected groups are:

Tumor
adjacent non-tumor
Normal

The Normal samples will later be excluded from the WGCNA network construction.

# 4. Standardize Sample Names

The expression matrix and metadata must use identical sample identifiers.

Here, suffixes beginning with an underscore are removed from both expression
sample names and metadata sample names.


```{r}

# Standardize expression sample names

colnames(expr) <- gsub(
  "_.*",
  "",
  colnames(expr)
)


# Standardize metadata sample names

rownames(meta) <- gsub(
  "_.*",
  "",
  rownames(meta)
)


# Display sample counts

cat("Number of expression samples:\n")
print(length(colnames(expr)))

cat("\nNumber of metadata samples:\n")
print(length(rownames(meta)))


# Determine common samples

cat("\nNumber of common samples:\n")

print(
  length(
    intersect(
      colnames(expr),
      rownames(meta)
    )
  )
)


```

# 5. Align Expression Matrix and Metadata

Only samples present in both the expression matrix and metadata are retained.

The ordering of samples is then made identical in both datasets.

This is essential because WGCNA assumes that every expression profile is
associated with the correct sample metadata.


```{r}

# Identify common samples

common_samples <- intersect(
  colnames(expr),
  rownames(meta)
)


# Keep only common samples

expr <- expr[
  ,
  common_samples,
  drop = FALSE
]

meta <- meta[
  common_samples,
  ,
  drop = FALSE
]


# Verify exact alignment

stopifnot(
  all(
    colnames(expr) == rownames(meta)
  )
)


# Display dimensions after alignment

cat("Aligned expression matrix dimensions:\n")
print(dim(expr))

cat("\nAligned metadata dimensions:\n")
print(dim(meta))


# Confirm successful alignment

cat("\nSample alignment successful:\n")

print(
  all(
    colnames(expr) == rownames(meta)
  )
)


```


# 6. Variance Filtering

Genes with very low expression variance provide limited information for
co-expression network construction.

Therefore, gene-wise variance is calculated across all available samples.

Importantly, the variance filtering step is performed before excluding the
Normal samples. This allows the Normal samples to contribute to identifying
variable genes.

Genes with variance greater than the 25th percentile are retained.



```{r}

# Convert expression data to a numeric matrix

expr_mat <- as.matrix(expr)


# Calculate variance for every gene

gene_var <- rowVars(
  expr_mat
)


# Calculate the 25th percentile variance cutoff

variance_cutoff <- quantile(
  gene_var,
  0.25
)


cat("25th percentile variance cutoff:\n")
print(variance_cutoff)


# Number of genes before filtering

cat("\nNumber of genes before filtering:\n")
print(
  length(gene_var)
)


# Retain genes with variance greater than the cutoff

expr_mat <- expr_mat[
  gene_var > variance_cutoff,
  ,
  drop = FALSE
]


# Convert filtered matrix back to data frame

expr <- as.data.frame(
  expr_mat
)


# Report number of retained genes

cat("\nNumber of genes after variance filtering:\n")
print(
  nrow(expr)
)


```

After this filtering step, approximately 75% of the original genes are
retained, depending on the distribution of gene variance.

# 7. Select Tumor and Adjacent Non-Tumor Samples

The three Normal samples are excluded from the WGCNA network construction.

Only Tumor and adjacent non-tumor samples are retained.


```{r}

# Select Tumor and adjacent non-tumor samples

keep_samples <- rownames(meta)[
  meta$tissue %in%
    c(
      "Tumor",
      "adjacent_non_tumor"
    )
]


# Create WGCNA expression matrix

expr_wgcna <- expr[
  ,
  keep_samples,
  drop = FALSE
]


# Create corresponding metadata

meta_sub <- meta[
  keep_samples,
  ,
  drop = FALSE
]


# Display number of WGCNA samples

cat("Number of WGCNA samples:\n")
print(
  nrow(meta_sub)
)


# Display tissue distribution

cat("\nWGCNA tissue groups:\n")

print(
  table(meta_sub$tissue)
)


# Display expression matrix dimensions

cat("\nExpression matrix dimensions after sample selection:\n")

print(
  dim(expr_wgcna)
)



```


The resulting dataset should contain the Tumor and adjacent non-tumor samples
only.

# 8. Prepare the WGCNA Expression Matrix

WGCNA requires the expression matrix to have:

Samples as rows
Genes as columns

The current expression matrix has genes as rows and samples as columns.
Therefore, the matrix is transposed.


```{r}

# Transpose expression matrix

datExpr <- t(
  expr_wgcna
)


# Convert to data frame

datExpr <- as.data.frame(
  datExpr
)


# Display dimensions

cat("WGCNA expression matrix dimensions:\n")

print(
  dim(datExpr)
)


cat("\nRows represent samples and columns represent genes.\n")


```


# 9. Align Metadata With Expression Matrix

The metadata is reordered so that its sample order exactly matches the rows of
the WGCNA expression matrix.

```{r}

# Reorder metadata according to expression matrix

meta_sub <- meta_sub[
  rownames(datExpr),
  ,
  drop = FALSE
]


# Confirm exact alignment

stopifnot(
  all(
    rownames(meta_sub) ==
      rownames(datExpr)
  )
)


cat("Metadata and expression matrix alignment successful:\n")

print(
  all(
    rownames(meta_sub) ==
      rownames(datExpr)
  )
)


```


# 10. Define the Tumor Trait

To evaluate the relationship between co-expression modules and pancreatic
cancer status, tissue status is converted into a numerical binary trait.

The coding is:

Tumor = 1
Adjacent non-tumor = 0

```{r}


# Create numerical tumor trait

traits_num <- data.frame(
  Tumor = ifelse(
    meta_sub$tissue == "Tumor",
    1,
    0
  )
)


# Assign sample names

rownames(traits_num) <- rownames(
  meta_sub
)


# Display trait distribution

cat("Tumor trait distribution:\n")

print(
  table(
    traits_num$Tumor
  )
)


cat("\nTrait coding:\n")
cat("Tumor = 1\n")
cat("Adjacent non-tumor = 0\n")

```

This binary trait will later be correlated with module eigengenes to determine
which modules are associated with pancreatic tumor status.

# 11. WGCNA Quality Control

The goodSamplesGenes() function checks whether samples and genes contain
excessive missing or problematic values.

Samples and genes that fail quality control can be removed before network
construction.


```{r}

# Perform WGCNA quality control

gsg <- goodSamplesGenes(
  datExpr,
  verbose = 3
)


# Display QC status

cat("All samples and genes passed initial WGCNA quality control:\n")

print(
  gsg$allOK
)


# Remove problematic samples or genes if necessary

if (!gsg$allOK) {
  
  datExpr <- datExpr[
    gsg$goodSamples,
    gsg$goodGenes,
    drop = FALSE
  ]
  
  
  meta_sub <- meta_sub[
    rownames(datExpr),
    ,
    drop = FALSE
  ]
  
  
  traits_num <- traits_num[
    rownames(datExpr),
    ,
    drop = FALSE
  ]
}


# Display final dimensions

cat("\nFinal dimensions after WGCNA QC:\n")

print(
  dim(datExpr)
)



```

# 12. Sample Clustering

Hierarchical clustering is used to examine the similarity between samples.

Samples that cluster far away from the main sample population may represent
potential outliers.

A horizontal reference line at height 150 is included based on the predefined
quality-control threshold.



```{r}

# Perform hierarchical clustering

sampleTree <- hclust(
  dist(datExpr),
  method = "average"
)


# Plot sample dendrogram

plot(
  sampleTree,
  main = "Sample Clustering",
  xlab = "",
  sub = "",
  cex = 0.6
)


# Add reference threshold

abline(
  h = 150,
  col = "red",
  lty = 2
)

png("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/wgcna/sample_tree.png", width = 1000, height = 600)
plot(sampleTree, main = "Sample clustering (Tumor vs Adjacent Non-Tumor)", sub = "", xlab = "", cex = 0.6)
abline(h = 150, col = "red")
dev.off()
```


The dendrogram is used to visually assess whether any samples are clear
outliers.

# 13. Principal Component Analysis

Principal component analysis (PCA) is performed to visualize the overall
structure of the expression data.

The first two principal components are plotted and colored according to tissue
type.


```{r}


# Perform PCA

pca <- prcomp(
  datExpr,
  scale. = TRUE
)


# Create PCA data frame

pca_df <- data.frame(
  pca$x[, 1:2],
  Tissue = meta_sub$tissue
)


# Plot PCA
png("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/wgcna/PCA_plot.png", width = 1000, height = 800)
ggplot(
  pca_df,
  aes(
    x = PC1,
    y = PC2,
    color = Tissue
  )
) +
  geom_point(
    size = 3
  ) +
  theme_minimal() +
  labs(
    title = "PCA of Tumor vs Adjacent Non-Tumor",
    x = paste0(
      "PC1 (",
      round(
        summary(pca)$importance[2, 1] * 100,
        2
      ),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(
        summary(pca)$importance[2, 2] * 100,
        2
      ),
      "%)"
    )
  )


dev.off()

```

Clear separation between Tumor and adjacent non-tumor samples would indicate
that tissue status is a major source of expression variation.

However, PCA is primarily a visualization tool and does not itself determine
whether a module is biologically significant.


# 14. Soft-Threshold Power Selection

WGCNA uses a soft-thresholding power to transform gene correlations into
network connection strengths.

Several candidate powers are evaluated to identify a power that produces a
network approximating scale-free topology while maintaining sufficient
connectivity.

A signed network is used so that positively correlated genes are emphasized
while negative correlations are treated differently.


```{r}

# Define candidate soft-thresholding powers

power <- c(
  1:10,
  seq(
    12,
    50,
    2
  )
)


# Evaluate candidate powers

sft <- pickSoftThreshold(
  datExpr,
  powerVector = power,
  networkType = "signed",
  verbose = 5
)


cat("Soft-threshold analysis completed.\n")
sft.data <- sft$fitIndices

cat("\nSoft-threshold fit indices:\n")

print(
  sft$fitIndices
)


```


# 15. Soft-Threshold Diagnostic Plot

Two diagnostic plots are generated:

Scale-free topology model fit
Mean connectivity

A reference line at R² = 0.8 is included to help evaluate the scale-free
topology criterion.
```{r}

# Plot scale-free topology
a1 <- ggplot(sft.data, aes(Power, SFT.R.sq, label = Power)) +
  geom_point() + geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.8, color = "red") +
  labs(x = "Soft threshold (Power)", y = "Scale-free topology model fit, signed R^2") +
  theme_classic()

a2 <- ggplot(sft.data, aes(Power, mean.k., label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(x = "Soft threshold (Power)", y = "Mean Connectivity") +
  theme_classic()
combined_plot <- arrangeGrob(a1, a2, nrow = 2)
a1
a2
ggsave("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/wgcna/soft_thresholding_plot_tumor_vs_adj.pdf", combined_plot, width = 8, height = 10)
# Save the first plot (Scale-free topology)
ggsave(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/wgcna/soft_thresholding_scale_free_tumor_vs_adj.png",
  plot = a1,
  width = 6, height = 5, dpi = 300
)

# Save the second plot (Mean connectivity)
ggsave(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/wgcna/soft_thresholding_mean_connectivity_tumor_vs_adj.png",
  plot = a2,
  width = 7, height = 5, dpi = 300
)


```



# 16. Select Soft-Threshold Power

Based on the previously evaluated scale-free topology and connectivity
results, a soft-thresholding power of 14 was selected for the network
construction.


```{r}

# Selected soft-thresholding power

soft_power <- 14


cat("Selected soft-thresholding power:\n")

print(
  soft_power
)


```

The selected power should be justified using the diagnostic plot and the
corresponding scale-free topology fit and connectivity values.

# 17. Construct the WGCNA Network

A signed weighted gene co-expression network is constructed using the selected
soft-thresholding power.

Genes with similar expression patterns across samples are grouped into
co-expression modules.

Modules with highly similar eigengene profiles are merged using a merge
cut-height of 0.25.


```{r}

# Construct weighted gene co-expression network

bwnet <- blockwiseModules(
  datExpr,
  power = soft_power,
  TOMType = "signed",
  mergeCutHeight = 0.25,
  numericLabels = FALSE,
  randomSeed = 1234,
  verbose = 3
)


cat("WGCNA network construction completed.\n")


```

# 18. Extract Module Information

The module color assigned to every gene and the module eigengenes are
extracted from the WGCNA results.

A module eigengene represents the major expression pattern of all genes within
a module.


```{r}

# Extract module assignments

moduleColors <- bwnet$colors


# Extract module eigengenes

MEs <- bwnet$MEs


# Display number of modules

cat("Number of detected modules:\n")

print(
  length(
    unique(moduleColors)
  )
)


# Display module sizes

cat("\nModule sizes:\n")

print(
  table(moduleColors)
)


```


# 19. Module Dendrogram

The module dendrogram visualizes the hierarchical organization of the detected
co-expression modules.

Each color represents a different co-expression module.



```{r}
# Number of genes in block 1
block1_n <- length(bwnet$dendrograms[[1]]$order)

# Extract colors corresponding to block 1
block1_unmerged <- bwnet$unmergedColors[1:block1_n]
block1_merged   <- bwnet$colors[1:block1_n]

png("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/wgcna/module_dendrogram_tumor_vs_adj.png", width = 1200, height = 800, res = 150)

# Plot block 1 dendrogram with unmerged and merged colors
plotDendroAndColors(
  bwnet$dendrograms[[1]],
  cbind(
    block1_unmerged,
    block1_merged
  ),
  c(
    "Unmerged",
    "Merged"
  ),
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05
)


dev.off()

```

# 20. Module-Trait Relationships

The relationship between each module eigengene and the Tumor trait is
calculated using Pearson correlation.

The resulting correlation indicates whether a module is positively or
negatively associated with tumor status.

P-values are calculated using the Student correlation significance test.


```{r}

# Calculate module-trait correlations

moduleTraitCor <- cor(
  MEs,
  traits_num,
  use = "p"
)


# Calculate corresponding p-values

moduleTraitPvalue <- corPvalueStudent(
  moduleTraitCor,
  nrow(datExpr)
)


# Display correlations

cat("Module-Trait Correlation:\n")

print(
  moduleTraitCor
)


# Display p-values

cat("\nModule-Trait P-values:\n")

print(
  moduleTraitPvalue
)


```


# 21. Prepare Module-Trait Heatmap Values

The correlation coefficient and corresponding p-value are combined into a
single text matrix.

The correlation is displayed first, followed by the p-value in parentheses.


```{r}

# Create heatmap annotation text

textMatrix <- paste(
  signif(
    moduleTraitCor,
    2
  ),
  "\n(",
  signif(
    moduleTraitPvalue,
    1
  ),
  ")",
  sep = ""
)


# Ensure dimensions match

dim(textMatrix) <- dim(
  moduleTraitCor
)


cat("Module-trait heatmap values prepared.\n")


```

# 22. Module-Trait Heatmap

The heatmap summarizes the relationship between WGCNA modules and tumor
status.

Positive correlations indicate higher module eigengene expression in Tumor
samples, whereas negative correlations indicate lower module eigengene
expression in Tumor samples.



```{r module_trait_heatmap,fig.width=16, fig.height=16,dpi=100,out.width="1600px",out.height="1600px"}


# ============================================================
# Module-Trait Relationship Heatmap
# ============================================================
# Save high-resolution heatmap
png(
  filename = "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/wgcna/module_trait_heatmap_tumor_vs_adj.png",
  width = 2700,
  height = 2400,
  res = 150
)

# IMPORTANT:
# Large left margin for complete module names
par(
  mar = c(6, 18, 5, 4),
  xpd = NA
)

# Module-Trait Relationship Heatmap
labeledHeatmap(
  Matrix = moduleTraitCor,

  xLabels = colnames(traits_num),

  # Complete module names
  yLabels = colnames(MEs),
  ySymbols = colnames(MEs),

  colorLabels = FALSE,
  colors = blueWhiteRed(50),

  textMatrix = textMatrix,

  # Prevent WGCNA from resetting the margins
  setStdMargins = FALSE,

  # Larger numbers inside the heatmap
  cex.text = 1.4,

  # Larger axis text
  cex.lab = 1.3,
  cex.axis = 1.2,

  zlim = c(-1, 1),

  main = "Module-Trait Relationships"
)

dev.off()

# Restore normal graphics settings
par(
  mar = c(5, 4, 4, 2) + 0.1,
  xpd = FALSE
)
```

# 23. Calculate Module Membership and Gene Significance

Two important measures are calculated for every gene:

Module Membership (MM)

Module Membership represents the correlation between an individual gene's
expression and the eigengene of its corresponding module.

A high absolute MM indicates that the gene is strongly representative of its
module.

Gene Significance (GS)

Gene Significance represents the correlation between an individual gene's
expression and the Tumor trait.

A high absolute GS indicates a strong association between the gene and tumor
status.



```{r}

# Store all gene names

all_genes <- colnames(
  datExpr
)


# Create empty result table

allModulesTable <- data.frame()


# Calculate MM and GS for every module

for (module in colnames(MEs)) {
  
  
  # Extract module color
  
  module_color <- gsub(
    "ME",
    "",
    module
  )
  
  
  # Identify genes belonging to this module
  
  inModule <- moduleColors ==
    module_color
  
  
  # Extract gene names
  
  modGenes <- all_genes[
    inModule
  ]
  
  
  # Calculate Module Membership
  
  MM <- cor(
    datExpr[
      ,
      inModule,
      drop = FALSE
    ],
    MEs[
      ,
      module
    ],
    use = "p"
  )
  
  
  # Calculate MM p-values
  
  MM_pval <- corPvalueStudent(
    MM,
    nrow(datExpr)
  )
  
  
  # Calculate Gene Significance
  
  GS <- cor(
    datExpr[
      ,
      inModule,
      drop = FALSE
    ],
    traits_num$Tumor,
    use = "p"
  )
  
  
  # Calculate GS p-values
  
  GS_pval <- corPvalueStudent(
    GS,
    nrow(datExpr)
  )
  
  
  # Create module-specific result table
  
  df <- data.frame(
    Gene = modGenes,
    Module = module_color,
    MM = as.vector(MM),
    MM_pval = as.vector(MM_pval),
    GS_Tumor = as.vector(GS),
    GS_pval = as.vector(GS_pval)
  )
  
  
  # Combine results
  
  allModulesTable <- rbind(
    allModulesTable,
    df
  )
  
}


# Display number of evaluated genes

cat("Total genes evaluated for MM and GS:\n")

print(
  nrow(allModulesTable)
)


# Display column names

cat("\nColumns in MM/GS table:\n")

print(
  colnames(allModulesTable)
)


```


# 24. Identify Significant Modules

Modules are considered significantly associated with Tumor status when both of
the following criteria are satisfied:

Absolute module-trait correlation > 0.5
Module-trait p-value < 0.05

These thresholds are used to identify modules with relatively strong and
statistically significant associations with pancreatic cancer status.


```{r}

# Identify significant modules

sigModules <- rownames(
  moduleTraitCor
)[
  abs(
    moduleTraitCor[
      ,
      "Tumor"
    ] 
  ) > 0.5 &
  moduleTraitPvalue[
    ,
    "Tumor"
  ] < 0.05
]


# Display significant modules

cat("Significant modules:\n")

print(
  sigModules
)


# Display number of significant modules

cat("\nNumber of significant modules:\n")

print(
  length(sigModules)
)


```


# 25. Add Significant Module Information to Gene Table

Each gene is labeled according to whether it belongs to one of the
significant modules identified above.




```{r}

# Identify genes belonging to significant modules

allModulesTable$SignificantModule <-
  allModulesTable$Module %in%
  gsub(
    "ME",
    "",
    sigModules
  )


# Count genes in significant modules

cat(
  "Number of genes belonging to significant modules:\n"
)

print(
  sum(
    allModulesTable$SignificantModule
  )
)


```

# 26. Extract Genes From Significant Modules

Only genes belonging to significant modules are retained for downstream hub
gene identification.



```{r}

# Extract genes from significant modules

significantModuleGenes <-
  allModulesTable %>%
  filter(
    SignificantModule == TRUE
  )


# Display number of genes

cat("Number of genes in significant modules:\n")

print(
  nrow(
    significantModuleGenes
  )
)


# Display genes per module

cat("\nGenes per significant module:\n")

print(
  table(
    significantModuleGenes$Module
  )
)


```

# 27. Identify Hub Genes

Hub genes are identified using three criteria:

The gene must belong to a significant module.
Absolute Module Membership (MM) must be greater than 0.8.
Absolute Gene Significance (GS) must be greater than 0.2.

These criteria identify genes that are strongly representative of their module
and are also associated with tumor status.



```{r}

# Identify hub genes

hubGenesTable <-
  allModulesTable %>%
  filter(
    SignificantModule == TRUE,
    abs(MM) > 0.8,
    abs(GS_Tumor) > 0.2
  )


# Display number of hub genes

cat("Number of hub genes:\n")

print(
  nrow(
    hubGenesTable
  )
)


# Display hub gene names

cat("\nHub genes:\n")

print(
  hubGenesTable$Gene
)



```

# 28. Sort Hub Genes

Hub genes are ranked according to:

Absolute Module Membership
Absolute Gene Significance

Genes with higher absolute MM and GS values are placed toward the top.


```{r}

# Sort hub genes

hubGenesTable <-
  hubGenesTable %>%
  arrange(
    desc(abs(MM)),
    desc(abs(GS_Tumor))
  )


# Display top 20 hub genes

cat("Top hub genes:\n")

print(
  head(
    hubGenesTable,
    20
  )
)


```

# 29. Save All Module Membership and Gene Significance Results

The complete MM and GS table is saved as a CSV file.

This file contains the module assignment, MM, MM p-value, GS, GS p-value, and
significant-module status for all analyzed genes.


```{r}

# Save all MM and GS results

write.csv(
  allModulesTable,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/wgcna/WGCNA_all_genes_MM_GS.csv",
  row.names = FALSE
)


cat("Saved:\n")
cat("WGCNA_all_genes_MM_GS.csv\n")


```


# 30. Save Hub Genes

The identified hub genes are saved for downstream analysis such as DEG
intersection, enrichment analysis, STRING network analysis, and biomarker
selection.


```{r}

# Save hub gene table

write.csv(
  hubGenesTable,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/wgcna/WGCNA_hub_genes.csv",
  row.names = FALSE
)


cat("Saved:\n")
cat("WGCNA_hub_genes.csv\n")


```

# 31. Save Significant Modules

The names of significant modules are saved as a text file.



```{r}

# Save significant module names

writeLines(
  sigModules,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/wgcna/WGCNA_significant_modules.txt"
)


cat("Saved:\n")
cat("WGCNA_significant_modules.txt\n")


```


# 32. Save Gene-Module Mapping

A gene-to-module mapping table is generated.

This file allows each gene to be linked to its corresponding WGCNA module.


```{r}

# Create gene-module mapping

gene_module_mapping <- data.frame(
  Gene = all_genes,
  Module = moduleColors
)


# Save mapping

write.csv(
  gene_module_mapping,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/wgcna/WGCNA_gene_module_mapping.csv",
  row.names = FALSE
)


cat("Saved:\n")
cat("WGCNA_gene_module_mapping.csv\n")


```

# 33. Final WGCNA Summary

The final section summarizes the major results of the WGCNA analysis,
including the number of genes, samples, modules, significant modules,
significant-module genes, and hub genes.


```{r}

cat("\n")
cat("============================================================\n")
cat("                 WGCNA FINAL SUMMARY\n")
cat("============================================================\n")


# Original expression dimensions

cat("\nOriginal expression dimensions:\n")

print(
  dim(expr_mat)
)


# Final WGCNA expression dimensions

cat("\nFinal WGCNA expression dimensions:\n")

print(
  dim(datExpr)
)


# Number of WGCNA samples

cat("\nNumber of WGCNA samples:\n")

print(
  nrow(datExpr)
)


# Number of WGCNA genes

cat("\nNumber of WGCNA genes:\n")

print(
  ncol(datExpr)
)


# Selected soft-threshold power

cat("\nSelected soft-threshold power:\n")

print(
  soft_power
)


# Number of modules

cat("\nNumber of modules:\n")

print(
  length(
    unique(moduleColors)
  )
)


# Significant modules

cat("\nSignificant modules:\n")

print(
  sigModules
)


# Number of significant modules

cat("\nNumber of significant modules:\n")

print(
  length(sigModules)
)


# Number of genes in significant modules

cat("\nNumber of genes in significant modules:\n")

print(
  nrow(
    significantModuleGenes
  )
)


# Number of hub genes

cat("\nNumber of hub genes:\n")

print(
  nrow(
    hubGenesTable
  )
)


cat("\n============================================================\n")
cat("        WGCNA analysis completed successfully.\n")
cat("============================================================\n")



```


# 34. Interpretation of the WGCNA Workflow

The WGCNA analysis follows a structured workflow in which highly variable genes
are first retained, followed by removal of Normal samples from the network
construction.

The resulting expression matrix is used to construct a signed weighted
co-expression network. Genes exhibiting similar expression patterns are
grouped into modules.

Module eigengenes are then correlated with the Tumor trait to identify modules
associated with pancreatic cancer status.

Within significant modules, genes with high Module Membership and Gene
Significance are identified as candidate hub genes.

These hub genes can subsequently be integrated with differential expression
results and other downstream analyses.

The overall analytical strategy can therefore be summarized as:

Normalized expression data

↓

Variance filtering

↓

Tumor + adjacent non-tumor samples

↓

WGCNA quality control

↓

Sample clustering + PCA

↓

Soft-threshold power selection

↓

Signed co-expression network

↓

Module identification

↓

Module-trait correlation

↓

Significant modules

↓

Module Membership + Gene Significance

↓

Hub genes

↓

Integration with DEG / enrichment / machine-learning analyses

