---
title: "Comparison of WGCNA Hub Genes and Differentially Expressed Genes"
author: "Iswarya L U"
date: "15-11-2025"
output: html_document
---


#necessary packages
```{r}
# ============================================================
# Venn Diagram: WGCNA Hub Genes vs DEGs
# ============================================================

library(tidyverse)
library(VennDiagram)
library(grid)

```

## Introduction


Differential expression analysis identifies genes whose expression differs significantly between tumor and adjacent non-tumor tissues.

WGCNA identifies groups of co-expressed genes associated with tumor status. Hub genes are highly connected genes within significant tumor-associated modules.

To identify potential candidate genes supported by both analyses, the WGCNA hub genes are compared with the significant differentially expressed genes.

The overlapping genes represent genes that are:

Differentially expressed between tumor and adjacent non-tumor tissues.
Highly connected within tumor-associated co-expression modules.

```{r}
# -----------------------
# Load data
# -----------------------
hub <- read.csv("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/wgcna/WGCNA_hub_genes.csv",stringsAsFactors = FALSE)
deg <- read.csv("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/differential_expression/csv_of_differentially_expressed_genes.csv",stringsAsFactors = FALSE)


```

The WGCNA results contain the identified hub genes.

The differential expression results contain all analyzed genes, including upregulated, downregulated, and non-significant genes.


## Extract WGCNA Hub Genes and Significant DEGs

```{r}
hub_genes <- hub %>%pull(Gene) %>%unique()

deg_genes <- deg %>%filter(status != "Not Significant") %>%pull(Gene) %>%  unique()

#Check Gene Identifiers

#Before calculating the overlap, it is important to verify that both datasets use the same type of gene identifier.

head(hub_genes)

head(deg_genes)

```

**pull(Gene)** extracts the gene identifiers from each dataset.

**unique()** removes duplicate gene identifiers.

For the DEG results, only genes classified as either **Upregulated or Downregulated** are retained.

Genes classified as **Not Significant** are excluded.

##Identify Overlapping Genes

```{r}
hub_deg_intersect <- intersect(hub_genes, deg_genes)

cat("Hub genes:", length(hub_genes), "\n")
cat("DEGs:", length(deg_genes), "\n")
cat("Overlap genes:", length(hub_deg_intersect), "\n")

```
The intersect() function identifies genes present in both the WGCNA hub gene list and the significant DEG list.

These overlapping genes represent potential candidate genes supported by both differential expression analysis and co-expression network analysis.

##Save Overlapping Genes

```{r}
overlap_genes <- data.frame(Gene = hub_deg_intersect)

write.csv(overlap_genes,"C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/Hub_DEG_overlap_genes.csv",row.names = FALSE)
```

##Venn Diagram of WGCNA Hub Genes and Significant DEGs

The Venn diagram provides a visual representation of the relationship between WGCNA hub genes and significant differentially expressed genes.

```{r,fig.width=12, fig.height=10, out.width='65%', fig.align='center'}
venn.plot <- venn.diagram(x = list(DEGs = deg_genes,WGCNA_Hub_Genes = hub_genes),filename = NULL,
fill = c("red","green"),
alpha = 0.5,
cex = 2,
cat.cex = 1.5,
margin = 0.1,
main = "Overlap of Significant DEGs and WGCNA Hub Genes"
)

grid.newpage()

grid.draw(venn.plot)
```



#SAVE THE VENN DIAGRAM

```{r}
png(
  
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/DEG_and_WGCNA_analysis/deg_wgcna_venn/WGCNA_DEG_Venn.png",width = 1200,height = 1000,res = 150)

grid.newpage()

grid.draw(venn.plot)

dev.off()
```

##view and save the integartion summary

```{r}
integration_summary <- data.frame(Category = c("WGCNA Hub Genes", "Significant DEGs","Overlapping Genes"),Number_of_Genes = c(length(hub_genes),length(deg_genes),length(hub_deg_intersect)))

integration_summary
#save the integration_summary

write.csv(integration_summary,"C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/DEG_and_WGCNA_analysis/deg_wgcna_venn/DEG_WGCNA_integration_summary.csv",row.names = FALSE
)

```


##Interpretation

The overlapping genes are particularly interesting because they satisfy two different criteria:

Significant Differential Expression
              +
High Connectivity in a Tumor-Associated WGCNA Module
              ↓
Potential Candidate Genes

This makes the DEG–WGCNA intersection a reasonable set of candidate genes for your subsequent biomarker discovery analyses.