---
title: "Functional Analysis of DEG ∩ WGCNA Overlap Genes"
author: "Iswarya L U"
date: "11-02-2026"
output: html_document
---
##necessary packages
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
library(DOSE)
library(dplyr)
library(pathview)

```

## Objective

The genes identified from the intersection of the differentially expressed genes (DEGs) and WGCNA-derived hub genes were subjected to functional enrichment analysis.

The purpose of this analysis was to determine whether the overlapping genes are significantly associated with particular:

Gene Ontology (GO) Biological Processes
Gene Ontology (GO) Cellular Components
Gene Ontology (GO) Molecular Functions
Disease Ontology (DO) terms
KEGG signaling pathways

This analysis helps determine the biological functions and pathways represented by the candidate genes identified through the integrated DEG and WGCNA analysis.

## 1.Load DEG ∩ WGCNA Overlap Genes

The genes common to the DEG and WGCNA analyses are loaded from the previously generated gene list.

```{r}
overlap_file <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/integration of deg and wgcna/Hub_DEG_overlap_genes.csv"

overlap_symbols <- read.table(
  overlap_file,
  header = TRUE,
  stringsAsFactors = FALSE
)

gene_symbols <- toupper(overlap_symbols$Gene)

```

he gene symbols are converted to uppercase to improve consistency with the human gene annotation database.

Input: DEG ∩ WGCNA overlap genes
Output: A standardized list of human gene symbols.

## 2.Convert Gene Symbols to Entrez Gene IDs

clusterProfiler enrichment functions commonly use Entrez Gene IDs for enrichment analysis. Therefore, the gene symbols are converted to Entrez IDs using org.Hs.eg.db.


```{r}
gene_df <- bitr(
  gene_symbols,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

entrez_ids <- unique(gene_df$ENTREZID)
```

Duplicate Entrez IDs are removed to ensure that each gene is represented only once in the enrichment analysis.

It is also useful to check how many genes were successfully mapped:



```{r}
cat("Input genes:", length(gene_symbols), "\n")
cat("Mapped Entrez IDs:", length(entrez_ids), "\n")

```

## 3.Define the Background Gene Universe

A background gene universe is used to reduce bias in enrichment analysis. Here, the background consists of human genes that can be mapped using the same annotation database.



```{r}
background_symbols <- keys(org.Hs.eg.db,keytype = "SYMBOL")

background_df <- bitr(
  background_symbols,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

background_ids <- unique(background_df$ENTREZID)

```

The background universe provides the reference set against which enrichment of the overlap genes is evaluated.

## 4.Gene Ontology Enrichment Analysis

Gene Ontology enrichment was performed separately for the three major GO categories:

Biological Process (BP) – biological processes in which the genes participate.
Cellular Component (CC) – cellular locations or structures associated with the genes.
Molecular Function (MF) – molecular activities performed by the gene products.

Benjamini–Hochberg (BH) correction is used to control the false discovery rate.




```{r}
#BP

ego_bp <- enrichGO(gene = entrez_ids,universe = background_ids,OrgDb = org.Hs.eg.db,ont = "BP", pAdjustMethod = "BH",qvalueCutoff = 0.05,readable = TRUE)

#CC

ego_cc <- enrichGO(
  gene = entrez_ids,
  universe = background_ids,
  OrgDb = org.Hs.eg.db,
  ont = "CC",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)




#MF

ego_mf <- enrichGO(
  gene = entrez_ids,
  universe = background_ids,
  OrgDb = org.Hs.eg.db,
  ont = "MF",
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE
)



```

#save the results in csv




```{r}
write.csv(as.data.frame(ego_bp),"C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/GO_BP_overlap.csv")

write.csv(as.data.frame(ego_cc),"C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/GO_CC_overlap.csv")

write.csv(as.data.frame(ego_mf),"C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/GO_MF_overlap.csv")

```



## 5.KEGG Pathway Enrichment

KEGG enrichment was performed to identify signaling and metabolic pathways significantly associated with the overlap genes.


```{r}
ekegg <- enrichKEGG(
  gene = entrez_ids,
  organism = "hsa",
  universe = background_ids,
  pvalueCutoff = 0.05
)

write.csv(as.data.frame(ekegg),"C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/KEGG_overlap.csv")

```

The enriched KEGG pathways can help identify biological signaling mechanisms potentially associated with the candidate genes.

## 6.Visualization of Enrichment Results

## 6.1 Combine GO Biological Process, Cellular Component and Molecular Function

The three GO enrichment results are combined into a single data frame to facilitate comparison and visualization.


```{r}
bp_df <- as.data.frame(ego_bp)
cc_df <- as.data.frame(ego_cc)
mf_df <- as.data.frame(ego_mf)

bp_df$ONTOLOGY <- "BP"
cc_df$ONTOLOGY <- "CC"
mf_df$ONTOLOGY <- "MF"

go_all <- rbind(
  bp_df,
  cc_df,
  mf_df
)

#The top seven enriched terms from each GO category are selected based on adjusted p-value.
go_all <- go_all %>%
  group_by(ONTOLOGY) %>%
  arrange(p.adjust) %>%
  dplyr::slice(1:7) %>%
  ungroup()

#The terms are ordered according to their appearance in the plot.
go_all$Description <- factor(
  go_all$Description,
  levels = rev(unique(go_all$Description))
)

```

## 7.2 GO Enrichment Bar Plot

A bar plot is generated to visualize the most significant GO terms across BP, CC and MF.

The bar length represents the number of overlap genes (Count) associated with each GO term, while the color represents the adjusted p-value.


```{r}
go_paper_plot <- ggplot(
  go_all,
  aes(
    x = Count,
    y = Description,
    fill = p.adjust
  )
) +
  geom_bar(stat = "identity") +
  facet_grid(
    ONTOLOGY ~ .,
    scales = "free_y",
    switch = "y"
  ) +
  scale_fill_gradient(
    low = "red",
    high = "blue"
  ) +
  labs(
    title = "GO Enrichment of Overlap Genes",
    x = "Count",
    y = NULL,
    fill = "P value"
  ) +
  theme_bw() +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(fill = "grey85"),
    strip.text.y.right = element_text(
      angle = 0,
      face = "bold"
    ),
    panel.spacing = unit(0.5, "lines")
  )


go_paper_plot

# to save 

ggsave("C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/GO_ALL_paper_style.png",go_paper_plot,width = 7,height = 8)

```



## 9. KEGG Pathway Visualization

The top 15 KEGG pathways are visualized using a dot plot.



```{r}
kegg_plot <- dotplot(
  ekegg,
  showCategory = 15
) +
  ggtitle("KEGG Pathway Enrichment – Overlap Genes") +
  theme_bw()

kegg_plot

ggsave( "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/KEGG_dotplot.png", kegg_plot, width = 8, height = 6)

kegg_bar <- barplot(
  ekegg,
  showCategory = 15
) +
  ggtitle("KEGG Pathway Enrichment – Overlap Genes") +
  theme_bw()


kegg_bar

ggsave( "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/KEGG_barplot.png",kegg_bar, width = 8, height = 6)
```


