---
title: "LASSO-Based Biomarker Selection for Pancreatic Cancer"
author: "Iswarya L U"
date: "2026-02-22"
output: html_document
---

```{r}
rm(list = ls())

options(stringsAsFactors = FALSE)

set.seed(123)

```

## Introduction

This analysis performs Least Absolute Shrinkage and Selection Operator (LASSO) regression to identify potential biomarkers that distinguish pancreatic tumor samples from adjacent non-tumor samples.

The input genes are obtained from the overlap between:

Differentially expressed genes (DEGs)
WGCNA hub genes

The workflow includes:

Loading normalized gene expression data.
Loading the overlapping DEG and WGCNA hub gene list.
Selecting tumor and adjacent non-tumor samples.
Preparing the machine learning feature matrix.
Performing LASSO logistic regression.
Selecting genes at lambda.min.
Extracting LASSO coefficients.
Generating coefficient path and cross-validation plots.

#Load libraries

```{r}
library(glmnet)
library(ggplot2)

```
The glmnet package is used to perform LASSO regression.

The ggplot2 package can be used for visualization.

## Load data


```{r}
expr <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/normalized_expression_matrix.csv",
  row.names = 1,
  check.names = FALSE
)

genes <- read.delim(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/integration of deg and wgcna/Hub_DEG_overlap_genes.csv",
  header = TRUE,
  stringsAsFactors = FALSE,
  sep = ","
)

meta <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/meta_clean.csv"
)

```
The analysis uses:

A normalized gene expression matrix.
A list of genes overlapping between DEG and WGCNA hub gene analysis.
Sample metadata containing tissue information.

#Prepare the Gene List

The first column of the gene file is extracted and duplicate genes are removed.
```{r}
gene_list <- unique(genes[[1]])

cat("Total genes loaded:", length(gene_list), "\n")

head(gene_list)

```
#Prepare the Expression Matrix

Only genes present in the overlap gene list are retained.

The expression matrix is then transposed because machine learning models require:

Samples as rows.
Genes as columns.




```{r}
expr_sub <- expr[
  rownames(expr) %in% gene_list,
  ,
  drop = FALSE
]

expr_t <- as.data.frame(
  t(expr_sub)
)

```

#Add Sample IDs

Sample names are standardized before merging with metadata.




```{r}
expr_t$SampleID <- gsub(
  "_.*",
  "",
  rownames(expr_t)
)

```

#Merge Expression Data and Metadata

The gene expression matrix is merged with the metadata using the sample identifier.


```{r}
merged_data <- merge(
  meta,
  expr_t,
  by = "SampleID"
)
```

#Select Tumor and Adjacent Non-Tumor Samples

Only the two groups relevant to the classification analysis are retained.

```{r}
merged_data <- merged_data[
  merged_data$tissue %in%
    c("Tumor", "adjacent_non_tumor"),
]

```

The classification task is therefore:

Tumor = 1
Adjacent non-tumor = 0

#Feature Selection and Dataset Preparation

Before building the LASSO model, the overlap between the requested gene list and genes available in the merged dataset is checked.


```{r}
valid_genes <- intersect(
  gene_list,
  colnames(merged_data)
)

cat("Genes requested:", length(gene_list), "\n")

cat("Genes found:", length(valid_genes), "\n")

```



#Create the machine learning dataset.


```{r}
ml_data <- merged_data[
  ,
  c("tissue", valid_genes),
  drop = FALSE
]

```


#Encode the Outcome Variable

The tissue type is converted into a binary outcome.


```{r}
ml_data$tissue <- factor(
  ml_data$tissue
)

y <- ifelse(
  ml_data$tissue == "Tumor",
  1,
  0
)

table(y)

```

#Create Predictor Matrix

The gene expression data are converted into a numerical matrix.


```{r}
x <- as.matrix(
  ml_data[, valid_genes]
)

x <- apply(
  x,
  2,
  as.numeric
)

```

#Scale Gene Expression

The gene expression values are standardized before LASSO analysis.



```{r}
x <- scale(x)

#missing values are removed 
complete_idx <- complete.cases(x)

x <- x[
  complete_idx,
  ,
  drop = FALSE
]

y <- y[complete_idx]

cat("Final samples:", nrow(x), "\n")

cat("Final genes:", ncol(x), "\n")
```
#LASSO Logistic Regression

LASSO regression is performed using the glmnet package.

The analysis uses:

Binomial logistic regression.
LASSO penalty (alpha = 1).
10-fold cross-validation.




```{r}
cvfit <- cv.glmnet(
  x,
  y,
  family = "binomial",
  alpha = 1,
  nfolds = 10,
  standardize = FALSE
)

fit <- glmnet(
  x,
  y,
  family = "binomial",
  alpha = 1,
  standardize = FALSE
)


cat(
  "Lambda.min:",
  cvfit$lambda.min,
  "\n"
)

cat(
  "Lambda.1se:",
  cvfit$lambda.1se,
  "\n"
)

```

#Interpretation of Lambda Values

lambda.min is the value of lambda that produces the minimum cross-validation error.

lambda.1se is the largest lambda within one standard error of the minimum cross-validation error.

The lambda.min model usually selects more genes, whereas the lambda.1se model generally produces a more parsimonious model.


#Extract LASSO-Selected Genes

Genes with non-zero coefficients at lambda.min are identified.



```{r}
coef_lasso <- coef(
  cvfit,
  s = "lambda.min"
)

lasso_genes <- rownames(coef_lasso)[
  coef_lasso[, 1] != 0
]

lasso_genes <- setdiff(
  lasso_genes,
  "(Intercept)"
)

cat(
  "Selected genes:",
  length(lasso_genes),
  "\n"
)

lasso_genes
```








```{r}

write.csv(
  lasso_genes,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/LASSO_selected_biomarkers.csv",
  row.names = FALSE
)

```


#Extract LASSO Coefficients

The coefficients of the selected genes are extracted.


```{r}
coef_df <- data.frame(
  Gene = lasso_genes,
  Coefficient = coef_lasso[
    lasso_genes,
    1
  ]
)

coef_df
#write.csv(coef_df, "~/work/r_project_ishu/ml/lasso/results/LASSO_coefficients.csv", row.names = FALSE)
```

A positive coefficient indicates that higher expression contributes toward classification as a tumor sample.

A negative coefficient indicates that higher expression contributes toward classification as an adjacent non-tumor sample.

#LASSO Coefficient Path

The coefficient path shows how the regression coefficients change as the regularization parameter lambda changes.


```{r}
coef_mat <- as.matrix(
  fit$beta
)

log_lambda <- log(
  fit$lambda
)

common_len <- min(
  ncol(coef_mat),
  length(log_lambda)
)

coef_mat <- coef_mat[
  ,
  1:common_len,
  drop = FALSE
]

log_lambda <- log_lambda[
  1:common_len
]

df_vals <- fit$df[
  1:common_len
]

idx <- order(
  log_lambda,
  decreasing = TRUE
)

matplot(
  log_lambda[idx],
  t(coef_mat[, idx]),
  type = "l",
  lty = 1,
  lwd = 2,
  xlab = "Log Lambda",
  ylab = "Coefficients",
  main = "LASSO Coefficient Profiles"
)

pretty_ticks <- pretty(
  log_lambda[idx]
)

tick_idx <- sapply(
  pretty_ticks,
  function(v) which.min(abs(log_lambda - v))
)

top_numbers <- df_vals[tick_idx]

axis(
  3,
  at = pretty_ticks,
  labels = top_numbers,
  tick = FALSE,
  line = -0.5,
  cex.axis = 1.2
)
```


#save coefficient path




```{r}
png(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/LASSO_Coefficient_Path.png",
  width = 2400,
  height = 1800,
  res = 300
)

matplot(
  log_lambda[idx],
  t(coef_mat[, idx]),
  type = "l",
  lty = 1,
  lwd = 2,
  xlab = "Log Lambda",
  ylab = "Coefficients",
  main = "LASSO Coefficient Profiles"
)

axis(
  3,
  at = pretty_ticks,
  labels = top_numbers,
  tick = FALSE,
  line = -0.5,
  cex.axis = 1.2
)

dev.off()

```


#LASSO Cross-Validation Plot

The cross-validation plot shows the binomial deviance across different lambda value.


```{r}
log_lambda_cv <- log(
  cvfit$lambda
)

idx_cv <- order(
  log_lambda_cv,
  decreasing = TRUE
)

plot(
  log_lambda_cv[idx_cv],
  cvfit$cvm[idx_cv],
  type = "b",
  pch = 16,
  col = "red",
  xlab = "Log(lambda)",
  ylab = "Binomial Deviance",
  main = "LASSO Cross-Validation"
)

arrows(
  log_lambda_cv[idx_cv],
  cvfit$cvm[idx_cv] -
    cvfit$cvsd[idx_cv],
  log_lambda_cv[idx_cv],
  cvfit$cvm[idx_cv] +
    cvfit$cvsd[idx_cv],
  angle = 90,
  code = 3,
  length = 0.03,
  col = "grey60"
)

abline(
  v = log(cvfit$lambda.min),
  lty = 2
)

abline(
  v = log(cvfit$lambda.1se),
  lty = 2
)

pretty_ticks_cv <- pretty(
  log_lambda_cv[idx_cv]
)

tick_idx_cv <- sapply(
  pretty_ticks_cv,
  function(v) which.min(
    abs(log_lambda_cv - v)
  )
)

top_numbers_cv <- cvfit$nzero[
  tick_idx_cv
]

axis(
  3,
  at = pretty_ticks_cv,
  labels = top_numbers_cv,
  tick = FALSE,
  line = -0.5,
  cex.axis = 1.2
)
```
#The two vertical lines represent:

lambda.min: minimum cross-validation error.
lambda.1se: a more conservative model within one standard error of the minimum.


#Save Cross-Validation Plot


```{r}
png(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/LASSO_CV_Plot.png",
  width = 2000,
  height = 1500,
  res = 300
)

plot(
  log_lambda_cv[idx_cv],
  cvfit$cvm[idx_cv],
  type = "b",
  pch = 16,
  col = "red",
  xlab = "Log(lambda)",
  ylab = "Binomial Deviance",
  main = "LASSO Cross-Validation"
)

arrows(
  log_lambda_cv[idx_cv],
  cvfit$cvm[idx_cv] -
    cvfit$cvsd[idx_cv],
  log_lambda_cv[idx_cv],
  cvfit$cvm[idx_cv] +
    cvfit$cvsd[idx_cv],
  angle = 90,
  code = 3,
  length = 0.03,
  col = "grey60"
)

abline(
  v = log(cvfit$lambda.min),
  lty = 2
)

abline(
  v = log(cvfit$lambda.1se),
  lty = 2
)

axis(
  3,
  at = pretty_ticks_cv,
  labels = top_numbers_cv,
  tick = FALSE,
  line = -0.5,
  cex.axis = 1.2
)

dev.off()
```


#LASSO Feature Count

This plot shows the number of genes with non-zero coefficients at different lambda values.



```{r}
plot(
  log_lambda_cv[idx_cv],
  cvfit$nzero[idx_cv],
  type = "l",
  lwd = 3,
  xlab = "Log(lambda)",
  ylab = "Number of Features",
  main = "LASSO Sparsity Profile"
)

points(
  log_lambda_cv[idx_cv],
  cvfit$nzero[idx_cv],
  pch = 16
)

abline(
  v = log(cvfit$lambda.min),
  lty = 2
)

abline(
  v = log(cvfit$lambda.1se),
  lty = 2
)

```

#Save Feature Count Plot


```{r}
png(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/LASSO_Feature_Count.png",
  width = 2000,
  height = 1500,
  res = 300
)

plot(
  log_lambda_cv[idx_cv],
  cvfit$nzero[idx_cv],
  type = "l",
  lwd = 3,
  xlab = "Log(lambda)",
  ylab = "Number of Features",
  main = "LASSO Sparsity Profile"
)

points(
  log_lambda_cv[idx_cv],
  cvfit$nzero[idx_cv],
  pch = 16
)

abline(
  v = log(cvfit$lambda.min),
  lty = 2
)

abline(
  v = log(cvfit$lambda.1se),
  lty = 2
)

dev.off()
```



#final result


```{r}
cat(
  "Number of input genes:",
  length(gene_list),
  "\n"
)

cat(
  "Number of genes available for LASSO:",
  length(valid_genes),
  "\n"
)

cat(
  "Number of samples:",
  nrow(x),
  "\n"
)

cat(
  "Number of LASSO-selected genes:",
  length(lasso_genes),
  "\n"
)

cat(
  "Lambda.min:",
  cvfit$lambda.min,
  "\n"
)

cat(
  "Lambda.1se:",
  cvfit$lambda.1se,
  "\n"
)

cat(
  "log(lambda.min):",
  log(cvfit$lambda.min),
  "\n"
)

cat(
  "log(lambda.1se):",
  log(cvfit$lambda.1se),
  "\n"
)

cat("Selected_biomarkers:",coef_df$Gene,"\n")
```
This analysis used LASSO logistic regression to select a reduced set of candidate biomarkers from the genes shared between the differential expression analysis and WGCNA hub gene analysis. The final biomarkers were selected based on non-zero coefficients at the cross-validated lambda.min value.
