---
title: "SVM-RFE-Based Biomarker Selection for Pancreatic Cancer"
author: "Iswarya L U"
date: "2026-02-22"
output: html_document
---

```{r}
rm(list = ls())

options(stringsAsFactors = FALSE)

set.seed(123)

```

A random seed is set to improve the reproducibility of the analysis.


## Introduction
This analysis performs **Support Vector Machine–Recursive Feature Elimination (SVM-RFE)** to identify a subset of candidate biomarkers capable of distinguishing pancreatic tumor samples from adjacent non-tumor samples.

The input genes consist of the overlap between:

- Differentially expressed genes (DEGs)
- WGCNA hub genes

SVM-RFE is used to evaluate different numbers of features and identify an optimal subset of genes for classification.

The workflow includes:

1. Loading the normalized gene expression matrix.
2. Loading the overlapping DEG and WGCNA hub gene list.
3. Selecting tumor and adjacent non-tumor samples.
4. Preparing and scaling the machine learning dataset.
5. Performing SVM-based recursive feature elimination.
6. Evaluating classification accuracy and cross-validation error.
7. Identifying the optimal number of features.
8. Extracting the selected biomarker genes.



# Load Required Libraries

```{r libraries}
library(caret)
library(e1071)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(doParallel)
```

The packages are used for:

- `caret`: machine learning workflow and recursive feature elimination.
- `e1071`: support vector machine implementation.
- `ggplot2`: visualization.
- `dplyr`: data manipulation.
- `gridExtra`: arranging graphical objects if needed.
- `doParallel`: parallel computing support.

# Configure Parallel Processing

```{r parallel-processing}
stopImplicitCluster()

registerDoSEQ()
```

This ensures that the analysis runs sequentially rather than using an implicit parallel backend.

# Load Data

```{r load-data}
expr <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/normalized_expression_matrix.csv",
  row.names = 1,
  check.names = FALSE
)

genes <- read.delim(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/integration of deg and wgcna/Hub_DEG_overlap_genes.csv",
  header = TRUE,
  stringsAsFactors = FALSE,
  sep = ","
)

meta <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/meta_clean.csv"
)
```

The analysis uses:

- A normalized gene expression matrix.
- A list of overlapping DEG and WGCNA hub genes.
- Sample metadata containing tissue information.

# Prepare the Gene List

The gene list is extracted and duplicate genes are removed.

```{r gene-list}
gene_list <- unique(genes$Gene)

cat(
  "Total genes:",
  length(gene_list),
  "\n"
)

head(gene_list)
```

# Filter the Expression Matrix

Only genes present in the overlap gene list are retained.

```{r filter-genes}
expr_filt <- expr[
  rownames(expr) %in% gene_list,
  ,
  drop = FALSE
]

cat(
  "Genes present in expression matrix:",
  nrow(expr_filt),
  "\n"
)
```

# Transpose the Expression Matrix

The original expression matrix contains:

- Genes as rows.
- Samples as columns.

Machine learning analysis requires:

- Samples as rows.
- Genes as columns.

Therefore, the expression matrix is transposed.

```{r transpose-expression}
expr_t <- as.data.frame(
  t(expr_filt)
)
```

# Prepare Sample IDs

Sample names are standardized to match the metadata.

```{r sample-id}
expr_t$SampleID <- gsub(
  "_.*",
  "",
  rownames(expr_t)
)
```

# Merge Expression Data and Metadata

The gene expression data are merged with the sample metadata.

```{r merge-data}
merged <- merge(
  meta,
  expr_t,
  by = "SampleID"
)
```

# Select Tumor and Adjacent Non-Tumor Samples

Only the two groups relevant to the classification analysis are retained.

```{r subset-samples}
merged <- merged[
  merged$tissue %in%
    c("Tumor", "adjacent_non_tumor"),
  ,
  drop = FALSE
]

merged$tissue <- factor(
  merged$tissue
)

table(merged$tissue)
```

The classification task consists of:

- Tumor
- Adjacent non-tumor

# Prepare the Machine Learning Dataset

The expression values of the candidate genes are extracted.

```{r ml-dataset}
valid_genes <- intersect(
  gene_list,
  colnames(merged)
)

cat(
  "Genes requested:",
  length(gene_list),
  "\n"
)

cat(
  "Genes available:",
  length(valid_genes),
  "\n"
)

ml_data <- merged[
  ,
  c("tissue", valid_genes),
  drop = FALSE
]
```

Using `valid_genes` rather than the complete `gene_list` ensures that the analysis does not fail if a gene from the input list is absent from the expression matrix.

# Prepare the Outcome Variable

The tissue labels are converted into a factor.

```{r outcome-variable}
y <- factor(
  ml_data$tissue
)

table(y)
```

The outcome variable contains two classes:

- `Tumor`
- `adjacent_non_tumor`

# Prepare the Predictor Matrix

The gene expression values are converted into a numeric data frame.

```{r predictor-matrix}
x <- as.data.frame(
  apply(
    ml_data[, valid_genes, drop = FALSE],
    2,
    as.numeric
  )
)
```

The predictor matrix is then standardized.

```{r scale-data}
x <- as.data.frame(
  scale(x)
)
```

# Remove Missing Values

Samples containing missing values are removed before machine learning analysis.

```{r remove-missing}
idx <- complete.cases(x)

x <- x[
  idx,
  ,
  drop = FALSE
]

y <- y[idx]

cat(
  "Final samples:",
  nrow(x),
  "\n"
)

cat(
  "Final genes:",
  ncol(x),
  "\n"
)
```

# SVM Recursive Feature Elimination

Recursive Feature Elimination (RFE) is performed using a Support Vector Machine with a radial basis function kernel.

Five-fold cross-validation is used to evaluate feature subsets.

```{r svm-rfe-control}
ctrl <- rfeControl(
  functions = caretFuncs,
  method = "cv",
  number = 5
)
```

The number of features evaluated ranges from one gene to a maximum of 50 genes.

```{r feature-sizes}
sizes <- seq(
  1,
  min(50, ncol(x)),
  by = 1
)
```

# Run SVM-RFE

```{r svm-rfe}
svmProfile <- rfe(
  x,
  y,
  sizes = sizes,
  rfeControl = ctrl,
  method = "svmRadial",
  tuneLength = 5
)
```

The SVM-RFE procedure evaluates different feature subset sizes and determines which subset provides the best classification performance.

# Extract SVM-RFE Results

The classification accuracy is extracted from the recursive feature elimination results.

```{r extract-results}
results <- svmProfile$results

results$CV_Error <- 1 - results$Accuracy

results
```

# Identify the Optimal Number of Features

The optimal number of features is determined based on the maximum cross-validation accuracy.

```{r optimal-features}
optimal_gene_number <- results$Variables[
  which.max(results$Accuracy)
]

cat(
  "Optimal gene number =",
  optimal_gene_number,
  "\n"
)
```

# Identify Optimal Accuracy and Error

The feature subset with the highest accuracy and the subset with the lowest cross-validation error are identified.

```{r optimal-accuracy-error}
opt_acc_idx <- which.max(
  results$Accuracy
)

opt_err_idx <- which.min(
  results$CV_Error
)

opt_features_acc <- results$Variables[
  opt_acc_idx
]

opt_features_err <- results$Variables[
  opt_err_idx
]

opt_acc <- round(
  results$Accuracy[opt_acc_idx],
  3
)

opt_err <- round(
  results$CV_Error[opt_err_idx],
  4
)

cat(
  "Optimal features based on accuracy =",
  opt_features_acc,
  "\n"
)

cat(
  "Optimal features based on error =",
  opt_features_err,
  "\n"
)

cat(
  "Maximum accuracy =",
  opt_acc,
  "\n"
)

cat(
  "Minimum CV error =",
  opt_err,
  "\n"
)
```

Since:

\[
\text{CV Error} = 1 - \text{Accuracy}
\]

the feature subset with maximum accuracy should correspond to the feature subset with minimum classification error.


# SVM-RFE Accuracy Plot

The following plot shows the classification accuracy obtained using different numbers of features.

```{r accuracy-plot, fig.width=8, fig.height=6}
par(mar = c(5, 5, 4, 2))

acc_range <- range(
  results$Accuracy
)

acc_pad <- diff(acc_range) * 0.08

plot(
  results$Variables,
  results$Accuracy,
  type = "l",
  col = "#4C84C4",
  lwd = 3,
  ylim = c(
    acc_range[1],
    acc_range[2] + acc_pad
  ),
  xlab = "Number of Features",
  ylab = "Accuracy",
  main = "SVM-RFE Classification Accuracy"
)

points(
  opt_features_acc,
  opt_acc,
  pch = 19
)

text(
  opt_features_acc,
  opt_acc + acc_pad * 0.5,
  labels = paste0(
    opt_features_acc,
    " – ",
    opt_acc
  ),
  col = "red",
  cex = 1.3
)
```

The highlighted point represents the number of features associated with the highest cross-validation accuracy.

# Save Accuracy Plot

```{r save-accuracy-plot}
png(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/SVM_RFE_Accuracy.png",
  width = 1600,
  height = 1200,
  res = 200
)

par(mar = c(5, 5, 4, 2))

plot(
  results$Variables,
  results$Accuracy,
  type = "l",
  col = "#4C84C4",
  lwd = 3,
  ylim = c(
    acc_range[1],
    acc_range[2] + acc_pad
  ),
  xlab = "Number of Features",
  ylab = "Accuracy",
  main = "SVM-RFE Classification Accuracy"
)

points(
  opt_features_acc,
  opt_acc,
  pch = 19
)

text(
  opt_features_acc,
  opt_acc + acc_pad * 0.5,
  labels = paste0(
    opt_features_acc,
    " – ",
    opt_acc
  ),
  col = "red",
  cex = 1.3
)

dev.off()
```

#  SVM-RFE Cross-Validation Error Plot

The cross-validation error is calculated as:

[CV\ Error = 1 - Accuracy]

```{r error-plot, fig.width=8, fig.height=6}
par(mar = c(5, 5, 4, 2))

err_range <- range(
  results$CV_Error
)

err_pad <- diff(err_range) * 0.08

plot(
  results$Variables,
  results$CV_Error,
  type = "l",
  col = "#4C84C4",
  lwd = 3,
  ylim = c(
    err_range[1] - err_pad,
    err_range[2] + err_pad
  ),
  xlab = "Number of Features",
  ylab = "5-Fold CV Error",
  main = "SVM-RFE Cross-Validation Error"
)

points(
  opt_features_err,
  opt_err,
  pch = 19
)

text(
  opt_features_err,
  opt_err - err_pad * 0.5,
  labels = paste0(
    opt_features_err,
    " – ",
    opt_err
  ),
  col = "red",
  cex = 1.3
)
```

# Save Error Plot

```{r save-error-plot}
png(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/SVM_RFE_Error.png",
  width = 1600,
  height = 1200,
  res = 200
)

par(mar = c(5, 5, 4, 2))

plot(
  results$Variables,
  results$CV_Error,
  type = "l",
  col = "#4C84C4",
  lwd = 3,
  ylim = c(
    err_range[1] - err_pad,
    err_range[2] + err_pad
  ),
  xlab = "Number of Features",
  ylab = "5-Fold CV Error",
  main = "SVM-RFE Cross-Validation Error"
)

points(
  opt_features_err,
  opt_err,
  pch = 19
)

text(
  opt_features_err,
  opt_err - err_pad * 0.5,
  labels = paste0(
    opt_features_err,
    " – ",
    opt_err
  ),
  col = "red",
  cex = 1.3
)

dev.off()
```

#Extract Selected Biomarker Genes

The optimal feature subset identified by SVM-RFE is extracted.

```{r extract-biomarkers}
svm_genes <- svmProfile$optVariables

cat(
  "Number of SVM-RFE selected biomarkers:",
  length(svm_genes),
  "\n"
)

svm_genes
```

#Save Selected Biomarker Genes

```{r save-biomarkers}
write.csv(
  data.frame(Gene = svm_genes),
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/SVM_RFE_biomarker_genes.csv",
  row.names = FALSE
)
```

# Final Summary

```{r final-summary}
cat(
  "Input genes:",
  length(gene_list),
  "\n"
)

cat(
  "Genes used for SVM-RFE:",
  ncol(x),
  "\n"
)

cat(
  "Samples used:",
  nrow(x),
  "\n"
)

cat(
  "Optimal number of features:",
  optimal_gene_number,
  "\n"
)

cat(
  "Maximum CV accuracy:",
  opt_acc,
  "\n"
)

cat(
  "Minimum CV error:",
  opt_err,
  "\n"
)

cat(
  "Selected biomarker genes:",
  length(svm_genes),
  "\n"
)
```

## Selected Biomarkers

```{r selected-biomarkers}
data.frame(
  Gene = svm_genes
)
```

# Conclusion

SVM-based recursive feature elimination was used to identify an optimal subset of genes capable of distinguishing pancreatic tumor samples from adjacent non-tumor samples. Starting with genes identified through the overlap between differential expression analysis and WGCNA hub gene analysis, SVM-RFE iteratively evaluated different feature subset sizes using five-fold cross-validation.

The final set of genes identified by `svmProfile$optVariables` represents the candidate biomarkers selected by the optimal SVM-RFE model.