---
title: "Overlap of Biomarker Genes Identified by Machine Learning Methods"
author: "Iswarya L U"
date: "2026-02-22"
output: html_document
---

# 1. Introduction

This analysis compares candidate biomarker genes identified using three different machine learning approaches:

- Support Vector Machine–Recursive Feature Elimination (SVM-RFE)
- Random Forest
- LASSO regression

The objective is to identify genes that are consistently selected across multiple machine learning methods.

The analysis includes:

1. Loading the biomarker gene lists.
2. Extracting gene names from each result.
3. Constructing a Venn diagram to visualize overlap.
4. Identifying genes common to all three machine learning methods.
5. Saving the common biomarker genes for downstream analysis.

# 2. Install and Load Required Packages

```{r setup}
if (!require(VennDiagram)) {
  install.packages("VennDiagram")
}

if (!require(ggvenn)) {
  install.packages("ggvenn")
}

library(VennDiagram)
library(ggvenn)

set.seed(123)
```

The packages are used for visualization of overlapping gene sets.

- `VennDiagram` creates traditional Venn diagrams.
- `ggvenn` creates a modern Venn diagram using the `ggplot2` visualization framework.

# 3. Load Machine Learning Gene Lists

The biomarker genes identified by SVM-RFE, Random Forest, and LASSO are loaded.

```{r load-gene-lists}
svm <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/SVM_RFE_biomarker_genes.csv",
  stringsAsFactors = FALSE
)

rf <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/RF_genes_importance_gt1.csv",
  stringsAsFactors = FALSE
)

lasso <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/LASSO_selected_biomarkers.csv",
  stringsAsFactors = FALSE
)
```

The three files contain the candidate biomarkers identified by the respective machine learning approaches.

# 4. Extract Gene Names

The gene names are extracted from the appropriate columns of each result file.

```{r extract-gene-names}
svm_genes <- unique(svm$Gene)

rf_genes <- unique(rf$Gene)

lasso_genes <- unique(lasso$x)
```

Duplicate gene names are removed using `unique()`.

# 5. Summarize the Number of Genes

```{r gene-counts}
cat(
  "SVM-RFE genes:",
  length(svm_genes),
  "\n"
)

cat(
  "Random Forest genes:",
  length(rf_genes),
  "\n"
)

cat(
  "LASSO genes:",
  length(lasso_genes),
  "\n"
)
```

# 6. Create Gene Sets

The three gene lists are combined into a named list for Venn diagram analysis.

```{r gene-sets}
gene_sets <- list(
  SVM_RFE = svm_genes,
  Random_Forest = rf_genes,
  LASSO = lasso_genes
)

gene_sets
```

# 7. Classic Venn Diagram

A traditional Venn diagram is generated to visualize the overlap between the three machine learning approaches.

```{r classic-venn}
venn.diagram(
  x = gene_sets,
  filename = "Biomarker_VennDiagram.png",
  fill = c(
    "red",
    "blue",
    "green"
  ),
  alpha = 0.5,
  cex = 1.5,
  cat.cex = 1.5,
  cat.pos = 0,
  main = "Overlap of Biomarker Genes from ML Methods"
)
```

The diagram shows:

- Genes uniquely identified by each method.
- Pairwise overlaps between methods.
- Genes shared by all three machine learning methods.

# 8. Publication-Style Venn Diagram

A modern Venn diagram is generated using `ggvenn`.

```{r modern-venn}
venn_plot <- ggvenn(
  gene_sets,
  fill_color = c(
    "red",
    "blue",
    "green"
  ),
  stroke_size = 0.5,
  set_name_size = 6
)

venn_plot
```

# 9. Save the Venn Diagram

```{r save-venn}
ggsave(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/Biomarker_VennDiagram_all_three_ml.png",
  venn_plot,
  width = 6,
  height = 6,
  dpi = 300
)
```

The figure is saved at 300 DPI, making it suitable for use in reports or presentations.

# 10. Identify Biomarkers Common to All Three Methods

The `Reduce(intersect, ...)` operation identifies genes that occur in all three gene sets.

```{r common-genes}
common_all <- Reduce(
  intersect,
  gene_sets
)

cat(
  "Number of common biomarkers:",
  length(common_all),
  "\n"
)

print(common_all)
```

These genes represent biomarkers that were independently selected by:

- SVM-RFE
- Random Forest
- LASSO

Therefore, they may represent more robust candidate biomarkers than genes selected by only one machine learning method.

# 11. Save Common Biomarker Genes

```{r save-common-genes}
write.csv(
  data.frame(Gene = common_all),
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/Common_Biomarker_Genes_All_3_ml.csv",
  row.names = FALSE
)
```

# 12. Final Summary

```{r final-summary}
cat(
  "====================================\n"
)

cat(
  "MACHINE LEARNING BIOMARKER OVERLAP\n"
)

cat(
  "====================================\n"
)

cat(
  "SVM-RFE genes:",
  length(svm_genes),
  "\n"
)

cat(
  "Random Forest genes:",
  length(rf_genes),
  "\n"
)

cat(
  "LASSO genes:",
  length(lasso_genes),
  "\n"
)

cat(
  "Common genes across all three:",
  length(common_all),
  "\n"
)

cat(
  "====================================\n"
)
```

# 13. Common Candidate Biomarkers

```{r display-common}
data.frame(
  Gene = common_all
)
```

# 14. Conclusion

Candidate biomarkers identified using SVM-RFE, Random Forest, and LASSO were compared to determine genes consistently selected across multiple machine learning approaches.

The intersection of the three gene sets was used to identify **common candidate biomarkers**. These genes represent candidates supported by multiple independent feature-selection strategies and can be subjected to further validation and functional analysis.

Potential downstream analyses include:

- Gene Ontology (GO) enrichment analysis
- KEGG pathway enrichment analysis
- Protein-protein interaction (PPI) network analysis
- Expression validation
- ROC curve analysis
- Survival analysis