---
title: External validation of identified biomarker efficiency using independent tumor
  dataset GSE62452
author: "Iswarya L U"
date: "2026-02-24"
output: html_document
---
# 1. Introduction

This analysis evaluates the diagnostic performance of the final biomarker genes identified from the pancreatic cancer discovery analysis using the independent **GSE62452** dataset.

GSE62452 contains gene-expression profiles from **69 pancreatic ductal adenocarcinoma (PDAC) tumor tissues and 61 adjacent non-tumor tissues**, generated using the Affymetrix Human Gene 1.0 ST Array (GPL6244).

The validation analysis consists of:

1. Extraction of the final biomarker genes.
2. Preparation of the independent validation dataset.
3. Comparison of gene expression between tumor and control tissues.
4. Evaluation of individual biomarker performance using ROC analysis.
5. Construction of a multi-gene logistic regression model.
6. Evaluation of the combined panel using ROC analysis.
7. Estimation of confidence intervals for the panel AUC.
8. Identification of the optimal classification threshold.
9. Generation of a risk-score dataset.
10. Visualization of individual-gene AUC values.

---

# 2. Dataset Information

## GSE62452

**GEO accession:** GSE62452

**Title:** Microarray gene-expression profiles of 69 pancreatic tumors and 61 adjacent non-tumor tissue from patients with pancreatic ductal adenocarcinoma.

**Organism:** *Homo sapiens*

**Platform:** GPL6244 – Affymetrix Human Gene 1.0 ST Array

**Samples:** 130 total samples

* 69 pancreatic tumor samples
* 61 adjacent non-tumor samples

The dataset was generated to investigate gene-expression differences between pancreatic tumor and adjacent non-tumor tissues in PDAC.

---

# 3. Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}

rm(list = ls())

graphics.off()

set.seed(123)

# Required packages
library(pROC)
library(ggpubr)
library(dplyr)
library(ggplot2)

```

A fixed random seed is used to improve reproducibility, particularly for bootstrap-based confidence interval estimation.

---

# 4. Define the Final Biomarker Genes

The final biomarker genes are read from the results file generated during the biomarker selection stage.

```{r biomarkers}

genes_read <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/Common_Biomarker_Genes_All_3_ml.csv",
  check.names = FALSE
)

final_genes <- genes_read$Gene

cat("Final biomarker genes:\n")
print(final_genes)

```

The genes contained in this file represent the final biomarker candidates selected from the discovery analysis.

---

# 5. Load Expression Matrix and Metadata

The normalized expression matrix and corresponding metadata for GSE62452 are loaded.

```{r load-data}

expr <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/GSE62452_RAW_normalized_expression_matrix_pancreas-tumor-val.csv",
  row.names = 1,
  check.names = FALSE
)

meta <- read.csv(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/GSE62452_RAW_metadata_pancreas_tissue_only--tumor-val.csv",
  row.names = 1,
  check.names = FALSE
)

cat("Expression matrix dimensions:\n")
print(dim(expr))

cat("\nMetadata dimensions:\n")
print(dim(meta))

```

The expression matrix contains genes as rows and samples as columns, whereas the metadata contains sample-level information.

---

# 6. Match Expression Data With Metadata

Only samples that are present in the metadata are retained from the expression matrix.

```{r match-samples}

# Check whether metadata samples are present in expression matrix
all(rownames(meta) %in% colnames(expr))

# Reorder expression matrix according to metadata
expr <- expr[, rownames(meta)]

# Confirm matching sample order
all(colnames(expr) == rownames(meta))

```

This step ensures that the expression measurements and phenotype information correspond to the same samples and are maintained in the same order.

---

# 7. Define Tumor and Control Groups

The original tissue annotation is converted into a binary classification variable:

* **Tumor** = pancreatic tumor tissue
* **Control** = adjacent non-tumor tissue

```{r define-groups}

meta$group <- factor(
  ifelse(meta$tissue == "tumor", "Tumor", "Control"),
  levels = c("Control", "Tumor")
)

# Display group counts
table(meta$group)

```

## Group Distribution

```{r group-distribution}

group_counts <- table(meta$group)

group_counts

```

The expected distribution is approximately:

* 61 Control samples
* 69 Tumor samples

---

# 8. Extract Biomarker Genes Present in the Validation Dataset

Only genes that are both present in the final biomarker list and available in the GSE62452 expression matrix are retained.

```{r extract-biomarkers}

common_genes <- intersect(
  final_genes,
  rownames(expr)
)

cat("Final genes identified:", length(final_genes), "\n")
cat("Genes available in GSE62452:", length(common_genes), "\n\n")

cat("Genes used for validation:\n")
print(common_genes)

```

The intersection step prevents errors caused by biomarker genes that may not be represented in the validation expression matrix.

---

# 9. Prepare the Biomarker Expression Dataset

The selected biomarker expression matrix is transposed so that samples are represented as rows and genes as columns.

```{r prepare-expression}

expr_biomarker <- t(
  expr[common_genes, , drop = FALSE]
)

expr_biomarker <- as.data.frame(
  expr_biomarker,
  check.names = FALSE
)

# Combine phenotype and expression data
df <- data.frame(
  group = meta$group,
  expr_biomarker,
  check.names = FALSE
)

# Display dimensions
dim(df)

# Preview data
head(df)

```

The resulting dataset contains one row per sample and one column for each biomarker gene.

---

# 10. Check for Missing Values

```{r missing-values}

cat("Total missing values in validation dataset:",
    sum(is.na(df)),
    "\n")

```

The presence of missing values should be assessed before statistical modeling.

---

# 11. Expression Distribution of Individual Biomarkers

Boxplots are generated to compare the expression levels of each biomarker between tumor and control tissues.

A Wilcoxon rank-sum test is used to assess the difference between the two groups.

```{r boxplots, fig.width=10, fig.height=7, message=FALSE, warning=FALSE}

for (g in common_genes) {

  p <- ggboxplot(
    df,
    x = "group",
    y = g,
    color = "group",
    add = "jitter",
    width = 0.65,
    size = 0.8
  ) +
    stat_compare_means(
      method = "wilcox.test"
    ) +
    labs(
      title = paste("GSE62452 Tumor vs Control Expression:", g),
      x = "Tissue Group",
      y = "Expression"
    ) +
    theme_bw(base_size = 14) +
    theme(
      plot.title = element_text(
        size = 16,
        face = "bold"
      ),
      axis.title = element_text(
        size = 14
      ),
      axis.text = element_text(
        size = 12
      ),
      legend.position = "none"
    )

  print(p)
}

```

These plots provide a visual assessment of the expression differences between tumor and adjacent non-tumor tissues.

---

# 12. Single-Gene ROC Analysis

Receiver operating characteristic (ROC) analysis is performed separately for each biomarker.

The area under the ROC curve (AUC) is used to quantify the ability of each individual gene to distinguish tumor from control samples.

An AUC closer to 1 indicates stronger discriminatory performance, whereas an AUC of 0.5 indicates performance equivalent to random classification.

```{r single-gene-roc, fig.width=9, fig.height=7}

roc_table <- data.frame(
  Gene = character(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)

for (g in common_genes) {

  roc_obj <- roc(
    response = df$group,
    predictor = df[, g],
    levels = c("Control", "Tumor"),
    direction = "<"
  )

  plot(
    roc_obj,
    lwd = 3,
    main = paste("ROC Curve:", g),
    print.auc = TRUE,
    print.auc.cex = 1.2
  )

  roc_table <- rbind(
    roc_table,
    data.frame(
      Gene = g,
      AUC = as.numeric(auc(roc_obj))
    )
  )
}

```

---

# 13. Single-Gene AUC Results

```{r single-gene-results}

cat("Single-gene AUCs for GSE62452:\n\n")

print(
  roc_table
)

```

The individual AUC values summarize the diagnostic discrimination achieved by each biomarker independently.

---

# 14. Save Single-Gene AUC Results

```{r save-single-auc}

write.csv(
  roc_table,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/GSE62452_SingleGene_AUC-.csv",
  row.names = FALSE
)

```

---

# 15. Multi-Gene Logistic Regression Model

A logistic regression model is constructed using the available biomarker genes simultaneously.

The model estimates the probability that each sample belongs to the tumor group based on the combined expression of the biomarker genes.

```{r logistic-model}

glm_model <- glm(
  group ~ .,
  data = df[, c("group", common_genes)],
  family = binomial
)

summary(glm_model)

```

---

# 16. Generate Predicted Tumor Probabilities

```{r predicted-probability}

pred <- predict(
  glm_model,
  type = "response"
)

head(pred)

```

The predicted probabilities represent the estimated probability of tumor classification for each sample.

These probabilities are subsequently used as the combined biomarker panel score for ROC analysis.

---

# 17. Multi-Gene ROC Analysis

The combined diagnostic performance of the biomarker panel is evaluated using ROC analysis.

```{r panel-roc}

roc_panel <- roc(
  response = df$group,
  predictor = pred,
  levels = c("Control", "Tumor"),
  direction = "<"
)

cat("Multi-gene panel AUC:",
    as.numeric(auc(roc_panel)),
    "\n")

```

---

# 18. Plot Multi-Gene ROC Curve

```{r panel-roc-plot, fig.width=9, fig.height=7}

plot(
  roc_panel,
  lwd = 3,
  main = "GSE62452 5-Gene Logistic Regression ROC",
  print.auc = TRUE,
  print.auc.cex = 1.2,
  print.auc.y = 0.2
)

```

The ROC curve evaluates the overall ability of the combined biomarker panel to discriminate between tumor and control tissues.

---

# 19. Save Multi-Gene ROC Plot

```{r save-panel-roc}

png(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/GSE62452_Multigene_AUC_plot-.png",
  width = 9,
  height = 7,
  units = "in",
  res = 300
)

plot(
  roc_panel,
  lwd = 3,
  main = "GSE62452 5-Gene Logistic ROC",
  print.auc = TRUE,
  print.auc.cex = 1.2,
  print.auc.y = 0.2
)

dev.off()

```

---

# 20. Extract and Save ROC Curve Coordinates

The sensitivity and specificity values corresponding to the ROC curve are saved for subsequent visualization or reporting.

```{r roc-coordinates}

roc_df <- data.frame(
  specificity = roc_panel$specificities,
  sensitivity = roc_panel$sensitivities
)

head(roc_df)

write.csv(
  roc_df,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/GSE62452_Logistic_ROC_curve.csv",
  row.names = FALSE
)

```

---

# 21. Generate Biomarker Risk Scores

The predicted probability generated by the logistic regression model is stored as the biomarker panel risk score.

```{r risk-score}

df$risk_score <- pred

head(
  df[, c("group", common_genes, "risk_score")]
)

```

The risk score represents the model-derived probability of tumor classification for each sample.

---

# 22. Save Risk Score Results

```{r save-risk-score}

write.csv(
  df,
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results/GSE62452_Biomarker_RiskScore.csv",
  row.names = FALSE
)

```

---

# 23. Confidence Interval for the Panel AUC

Two confidence interval estimates are obtained for the combined panel AUC:

1. The default DeLong method.
2. A bootstrap-based confidence interval using 2,000 bootstrap replicates.

```{r auc-confidence-interval}

# DeLong confidence interval
ci_auc <- ci.auc(
  roc_panel
)

# Bootstrap confidence interval
ci_auc_boot <- ci.auc(
  roc_panel,
  method = "bootstrap",
  boot.n = 2000
)

cat("AUC:\n")
print(auc(roc_panel))

cat("\nDeLong 95% CI:\n")
print(ci_auc)

cat("\nBootstrap 95% CI:\n")
print(ci_auc_boot)

```

The confidence intervals provide an estimate of the uncertainty surrounding the observed panel AUC.

---

# 24. Optimal Classification Threshold

The threshold that provides the best combination of sensitivity and specificity is identified using the `"best"` criterion implemented in `pROC`.

```{r optimal-threshold}

coords_best <- coords(
  roc_panel,
  "best",
  ret = c(
    "threshold",
    "sensitivity",
    "specificity",
    "ppv",
    "npv"
  )
)

cat("Best classification threshold and diagnostic metrics:\n\n")

print(coords_best)

```

The resulting threshold can be used to classify samples according to the predicted probability generated by the logistic regression panel.

---

# 25. Arrange Biomarkers According to AUC

The individual biomarkers are ranked according to their AUC values for visualization.

```{r auc-ranking}

roc_table$Gene <- factor(
  roc_table$Gene,
  levels = roc_table$Gene[
    order(roc_table$AUC)
  ]
)

roc_table

```

---

# 26. Single-Gene AUC Bar Plot

A horizontal bar plot is generated to compare the diagnostic performance of the individual biomarkers.

```{r auc-barplot, fig.width=10, fig.height=7}

p_auc <- ggplot(
  roc_table,
  aes(
    x = Gene,
    y = AUC,
    fill = Gene
  )
) +

  geom_bar(
    stat = "identity",
    width = 0.6
  ) +

  geom_text(
    aes(
      label = round(AUC, 3)
    ),
    hjust = -0.1,
    size = 4
  ) +

  coord_flip() +

  scale_y_continuous(
    limits = c(0, 1),
    expand = expansion(
      mult = c(0, 0.08)
    )
  ) +

  labs(
    title = "Single-Gene AUCs in GSE62452",
    x = "Biomarker Gene",
    y = "Area Under the ROC Curve (AUC)"
  ) +

  theme_bw(
    base_size = 14
  ) +

  theme(
    legend.position = "none",
    plot.title = element_text(
      size = 16,
      face = "bold"
    ),
    axis.title = element_text(
      size = 14
    ),
    axis.text = element_text(
      size = 12
    )
  )

print(p_auc)

```

---

# 27. Save Single-Gene AUC Plot

```{r save-auc-barplot}

ggsave(
  "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures/GSE62452_SingleGene_AUC_plot.png",
  plot = p_auc,
  width = 9,
  height = 7,
  dpi = 300
)

```

---

# 28. Final Validation Summary

```{r final-summary}

panel_auc <- as.numeric(
  auc(roc_panel)
)

cat("============================================\n")
cat("GSE62452 BIOMARKER VALIDATION SUMMARY\n")
cat("============================================\n\n")

cat("Total samples:", nrow(df), "\n")
cat("Control samples:",
    sum(df$group == "Control"),
    "\n")

cat("Tumor samples:",
    sum(df$group == "Tumor"),
    "\n\n")

cat("Genes evaluated:",
    length(common_genes),
    "\n")

cat("Genes:\n")
print(common_genes)

cat("\nSingle-gene AUCs:\n")
print(roc_table)

cat("\nCombined panel AUC:",
    round(panel_auc, 3),
    "\n")

cat("\nDeLong 95% CI:\n")
print(ci_auc)

cat("\nBootstrap 95% CI:\n")
print(ci_auc_boot)

cat("\nOptimal threshold metrics:\n")
print(coords_best)

cat("\n============================================\n")

```

# 29. Interpretation

The GSE62452 dataset provides an independent tissue-level validation cohort containing pancreatic tumor and adjacent non-tumor samples.

The individual biomarkers are first evaluated independently using ROC analysis. Their AUC values indicate the ability of each gene to discriminate between tumor and control tissues.

A logistic regression model is then used to combine the expression information from the biomarker genes. The predicted probabilities from this model are treated as a combined biomarker score and evaluated using ROC analysis.

The resulting panel AUC, confidence intervals, sensitivity, specificity, PPV, NPV, and optimal threshold provide a quantitative assessment of the diagnostic performance of the combined biomarker panel.

The GSE62452 analysis therefore serves as an **external tissue validation step** for the biomarker candidates identified during the discovery phase.

---

# 30. Files Generated

The analysis generates the following output files:

### Results

* `GSE62452_SingleGene_AUC.csv`
* `GSE62452_Logistic_ROC_curve.csv`
* `GSE62452_Biomarker_RiskScore.csv`

### Figures

* `GSE62452_Multigene_AUC_plot.png`
* `GSE62452_SingleGene_AUC_plot.png`

