---

title: "External Validation of Pancreatic Cancer Biomarkers Using peripheral blood transcriptomic dataset GSE15932"
author: "Iswarya L U"
date: "24-02-2026"
output:
html_document:
toc: true
toc_float: true
theme: readable
---------------

# GSE15932 External Validation

## 1. Introduction

This analysis evaluates the diagnostic performance of a selected pancreatic cancer biomarker panel using the independent blood-based microarray dataset **GSE15932**.

The selected biomarkers are:

* **AHNAK2**
* **TRIM29**
* **SLC6A14**
* **ITGB4**
* **CTSE**

The analysis includes:

1. Loading normalized gene-expression data and metadata.
2. Restricting the dataset to pancreatic cancer and healthy control samples.
3. Extracting expression values for the selected biomarkers.
4. Fitting a multi-gene logistic regression model.
5. Generating predicted risk scores.
6. Evaluating the multi-gene panel using ROC analysis.
7. Calculating AUC and 95% confidence intervals.
8. Evaluating each biomarker individually.
9. Comparing single-gene and multi-gene ROC curves.
10. Identifying the optimal classification threshold using the Youden index.
11. Constructing a confusion matrix.
12. Saving risk scores and ROC coordinates.

---

# 2. Dataset Information

## GSE15932

**GEO accession:** GSE15932

**Title:** Blood biomarkers of pancreatic cancer associated diabetes identified by peripheral blood-based gene expression profiles

**Organism:** *Homo sapiens*

**Experiment type:** Expression profiling by array

**Platform:** GPL570 – Affymetrix Human Genome U133 Plus 2.0 Array

The original study included 32 peripheral blood samples:

* 8 pancreatic cancer with diabetes
* 8 pancreatic cancer without diabetes
* 8 diabetes mellitus
* 8 healthy controls

For the present external validation analysis, only:

* **Pancreatic cancer without diabetes (PC only)**
* **Healthy controls**

are retained.

The pancreatic cancer-associated diabetes and diabetes-only groups are excluded from this analysis.

---

# 3. Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}

rm(list = ls())
graphics.off()

set.seed(123)

# Load libraries
library(pROC)
library(ggpubr)

```

A random seed is set to ensure reproducibility of analyses involving random resampling, such as bootstrap confidence intervals.

---

# 4. Define Final Biomarker Panel

The final biomarker panel consists of five genes selected from the pancreatic cancer biomarker discovery analysis.

```{r define-biomarkers}

final_genes <- c(
  "AHNAK2",
  "TRIM29",
  "SLC6A14",
  "ITGB4",
  "CTSE"
)

final_genes

```

---

# 5. Define File Paths

The normalized expression matrix and metadata are loaded from the project directory.

```{r file-paths}

expr_file <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/normalized_expression_matrix_GSE15932.csv"

meta_file <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/data/metadata_GSE15932_clean.csv"

figure_dir <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/figures"

result_dir <- "C:/Users/iswar/Downloads/pancreatic_cancer_biomarker_discovery/results"

single_gene_roc_dir <- file.path(
  figure_dir,
  "single_gene_ROC"
)

# Create output directories if they do not already exist

dir.create(
  figure_dir,
  showWarnings = FALSE,
  recursive = TRUE
)

dir.create(
  result_dir,
  showWarnings = FALSE,
  recursive = TRUE
)

dir.create(
  single_gene_roc_dir,
  showWarnings = FALSE,
  recursive = TRUE
)

```

---

# 6. Load Expression and Metadata

```{r load-data}

expr <- read.csv(
  expr_file,
  row.names = 1,
  check.names = FALSE
)

meta <- read.csv(
  meta_file,
  stringsAsFactors = FALSE
)

```

## Inspect Expression Data

```{r inspect-expression}

cat("Expression matrix dimensions:\n")

print(dim(expr))

cat("\nFirst few genes:\n")

print(head(rownames(expr)))

```

## Inspect Metadata

```{r inspect-metadata}

cat("Metadata dimensions:\n")

print(dim(meta))

cat("\nOriginal group distribution:\n")

print(table(meta$Group))

```

---

# 7. Recode Study Groups

The original metadata contains four groups.

For this external validation analysis, only **PC only** and **Healthy** samples are retained.

```{r recode-groups}

meta$group <- NA

meta$group[
  meta$Group == "PC only"
] <- "Tumor"

meta$group[
  meta$Group == "Healthy"
] <- "Control"

```

---

# 8. Subset Pancreatic Cancer and Healthy Controls

Samples belonging to:

* PC + Diabetes
* Diabetes only

are excluded.

```{r subset-groups}

meta <- meta[
  !is.na(meta$group),
]

meta$group <- factor(
  meta$group,
  levels = c("Control", "Tumor")
)

cat("Filtered group distribution:\n")

print(table(meta$group))

```

The resulting dataset therefore represents a binary classification problem:

**Control vs Tumor**

---

# 9. Extract Biomarker Expression

Only genes present in both the predefined biomarker panel and the expression matrix are retained.

```{r extract-genes}

common_genes <- intersect(
  final_genes,
  rownames(expr)
)

cat(
  "Genes used in the analysis:\n"
)

print(common_genes)

```

Check whether any biomarker is missing:

```{r check-missing-genes}

missing_genes <- setdiff(
  final_genes,
  common_genes
)

cat("Missing genes:\n")

print(missing_genes)

```

The expression matrix is transposed so that samples are represented as rows and genes as columns.

```{r transpose-expression}

expr_t <- as.data.frame(
  t(expr[common_genes, ])
)

expr_t$SampleID <- rownames(expr_t)

```

---

# 10. Clean Sample IDs and Merge Data

Sample identifiers are cleaned before merging the metadata and expression matrix.

```{r merge-data}

meta$SampleID <- trimws(
  as.character(meta$SampleID)
)

expr_t$SampleID <- trimws(
  as.character(expr_t$SampleID)
)

df <- merge(
  meta,
  expr_t,
  by = "SampleID"
)

df$group <- factor(
  df$group,
  levels = c("Control", "Tumor")
)

```

## Final Dataset Size

```{r final-dataset}

cat(
  "Final dataset size:",
  nrow(df),
  "samples\n"
)

cat("\nFinal group distribution:\n")

print(table(df$group))

```

---

# 11. Logistic Regression Model

A multivariable logistic regression model is fitted using the selected biomarker genes.

The model estimates the probability that a sample belongs to the **Tumor** group.

```{r logistic-regression}

glm_model <- glm(
  group ~ .,
  data = df[, c("group", common_genes)],
  family = binomial(link = "logit")
)

```

## Model Summary

```{r model-summary}

cat("\nLogistic regression model summary:\n\n")

print(summary(glm_model))

```

---

# 12. Generate Predicted Risk Scores

Predicted probabilities are generated from the logistic regression model.

```{r predicted-risk}

pred <- predict(
  glm_model,
  type = "response"
)

```

## Prediction Summary

```{r prediction-summary}

cat("\nPrediction summary:\n\n")

print(summary(pred))

```

The predicted probability represents the estimated probability that each sample belongs to the tumor group.

---

# 13. Multi-Gene ROC Analysis

The predicted risk scores from the logistic regression model are evaluated using receiver operating characteristic (ROC) analysis.

```{r multi-gene-roc}

roc_panel <- pROC::roc(
  response = df$group,
  predictor = pred
)

```

---

# 14. Multi-Gene AUC

The area under the ROC curve (AUC) is calculated to evaluate the discriminatory performance of the multi-gene panel.

```{r multi-gene-auc}

cat("\nMulti-gene AUC:\n")

print(
  auc(roc_panel)
)

```

An AUC closer to 1 indicates stronger discrimination between pancreatic cancer and healthy control samples.

---

# 15. DeLong 95% Confidence Interval

The 95% confidence interval for the AUC is calculated using the DeLong method.

```{r delong-ci}

cat("\nDeLong 95% CI:\n")

print(
  ci.auc(roc_panel)
)

```

---

# 16. Bootstrap 95% Confidence Interval

A bootstrap confidence interval is also calculated using 2000 resamples.

```{r bootstrap-ci}

cat("\nBootstrap 95% CI using 2000 resamples:\n")

print(
  ci.auc(
    roc_panel,
    method = "bootstrap",
    boot.n = 2000
  )
)

```

---

# 17. Single-Gene ROC Analysis

Each biomarker is evaluated individually using ROC analysis.

```{r single-gene-roc}

single_auc <- data.frame(
  Gene = character(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)

for (gene in common_genes) {

  roc_single <- roc(
    df$group,
    df[[gene]]
  )

  auc_val <- auc(
    roc_single
  )

  single_auc <- rbind(
    single_auc,
    data.frame(
      Gene = gene,
      AUC = as.numeric(auc_val)
    )
  )

  # Save individual ROC curve

  png(
    file.path(
      single_gene_roc_dir,
      paste0(gene, "_ROC.png")
    ),
    width = 900,
    height = 750
  )

  plot(
    roc_single,
    col = "red",
    lwd = 4,
    main = paste(
      gene,
      "ROC"
    ),
    cex.main = 1.5,
    print.auc = TRUE,
    print.auc.cex = 1.4
  )

  text(
    0.6,
    0.2,
    labels = paste(
      "AUC =",
      round(auc_val, 3)
    ),
    cex = 1.2
  )

  dev.off()

}

```

---

# 18. Rank Individual Biomarkers by AUC

```{r rank-single-gene}

single_auc <- single_auc[
  order(
    -single_auc$AUC
  ),
]

print(single_auc)

```

---

# 19. Save Single-Gene AUC Results

```{r save-single-gene-auc}

write.csv(
  single_auc,
  file.path(
    result_dir,
    "GSE15932_single_gene_AUC.csv"
  ),
  row.names = FALSE
)

```

---

# 20. Plot Multi-Gene ROC Curve

The multi-gene logistic regression ROC curve is plotted with an enlarged canvas for better visualization.

```{r multi-gene-roc-plot, fig.width=10, fig.height=8}

png(
  file.path(
    figure_dir,
    "GSE15932_multi_gene_ROC.png"
  ),
  width = 1200,
  height = 1000
)

plot(
  roc_panel,
  col = "blue",
  lwd = 5,
  main = "GSE15932 Multi-Gene Logistic ROC",
  cex.main = 1.5,
  print.auc = TRUE,
  print.auc.cex = 1.5,
  print.auc.y = 0.25
)

# Diagonal reference line

abline(
  a = 0,
  b = 1,
  lty = 2,
  col = "gray"
)

# Calculate DeLong confidence interval

ci_val <- ci.auc(
  roc_panel
)

text(
  0.6,
  0.1,
  labels = paste(
    "95% CI:",
    round(ci_val[1], 3),
    "-",
    round(ci_val[3], 3)
  ),
  cex = 1.2
)

dev.off()

```

---

# 21. Display Multi-Gene ROC Curve

```{r display-multi-gene-roc}

plot(
  roc_panel,
  col = "blue",
  lwd = 4,
  main = "GSE15932 Multi-Gene Logistic ROC",
  print.auc = TRUE,
  print.auc.cex = 1.3
)

abline(
  a = 0,
  b = 1,
  lty = 2,
  col = "gray"
)

```

---

# 22. Overlay Multi-Gene and Single-Gene ROC Curves

The multi-gene logistic regression model is compared with the individual biomarkers.

```{r roc-overlay}

png(
  file.path(
    figure_dir,
    "GSE15932_ROC_overlay.png"
  ),
  width = 1200,
  height = 1000
)

# Multi-gene ROC

plot(
  roc_panel,
  col = "black",
  lwd = 5,
  main = "Multi vs Single Gene ROC",
  print.auc = TRUE,
  print.auc.cex = 1.4,
  print.auc.y = 0.25
)

# Prepare legend

legend_labels <- c(
  paste0(
    "Multi-gene (AUC = ",
    round(
      auc(roc_panel),
      3
    ),
    ")"
  )
)

# Colors for individual genes

cols <- c(
  "red",
  "blue",
  "green",
  "purple",
  "orange"
)

i <- 1

for (gene in common_genes) {

  roc_single <- roc(
    df$group,
    df[[gene]]
  )

  auc_val <- auc(
    roc_single
  )

  # Add individual ROC curve

  plot(
    roc_single,
    add = TRUE,
    col = cols[i],
    lwd = 2
  )

  # Add AUC to legend

  legend_labels <- c(
    legend_labels,
    paste0(
      gene,
      " (AUC = ",
      round(
        auc_val,
        3
      ),
      ")"
    )
  )

  i <- i + 1

}

# Add legend

legend(
  "bottomright",
  legend = legend_labels,
  col = c(
    "black",
    cols
  ),
  lwd = c(
    5,
    rep(
      2,
      length(common_genes)
    )
  ),
  cex = 1.4,
  pt.cex = 1.6,
  y.intersp = 1.5,
  x.intersp = 1.3,
  seg.len = 3.5,
  box.lwd = 2,
  inset = 0.02
)

dev.off()

```

---

# 23. Display ROC Overlay

```{r display-roc-overlay, fig.width=11, fig.height=9}

plot(
  roc_panel,
  col = "black",
  lwd = 5,
  main = "GSE15932: Multi-Gene vs Single-Gene ROC",
  print.auc = TRUE,
  print.auc.cex = 1.3,
  print.auc.y = 0.25
)

legend_labels <- c(
  paste0(
    "Multi-gene (AUC = ",
    round(
      auc(roc_panel),
      3
    ),
    ")"
  )
)

cols <- c(
  "red",
  "blue",
  "green",
  "purple",
  "orange"
)

i <- 1

for (gene in common_genes) {

  roc_single <- roc(
    df$group,
    df[[gene]]
  )

  auc_val <- auc(
    roc_single
  )

  plot(
    roc_single,
    add = TRUE,
    col = cols[i],
    lwd = 2
  )

  legend_labels <- c(
    legend_labels,
    paste0(
      gene,
      " (AUC = ",
      round(
        auc_val,
        3
      ),
      ")"
    )
  )

  i <- i + 1

}

legend(
  "bottomright",
  legend = legend_labels,
  col = c(
    "black",
    cols
  ),
  lwd = c(
    5,
    rep(
      2,
      length(common_genes)
    )
  ),
  cex = 1.2,
  y.intersp = 1.3,
  seg.len = 3
)

```

---

# 24. Determine Optimal Cutoff Using the Youden Index

The optimal classification threshold is identified using the Youden index.

The Youden index maximizes:

**Sensitivity + Specificity - 1**

```{r youden-cutoff}

opt_cut <- coords(
  roc_panel,
  "best",
  ret = c(
    "threshold",
    "sensitivity",
    "specificity"
  ),
  best.method = "youden"
)

cat("\nOptimal cutoff using Youden index:\n\n")

print(opt_cut)

```

---

# 25. Classification Using the Optimal Cutoff

The first optimal threshold returned by the ROC analysis is used for classification.

```{r classification}

threshold <- opt_cut$threshold[1]

pred_class <- ifelse(
  pred >= threshold,
  "Tumor",
  "Control"
)

pred_class <- factor(
  pred_class,
  levels = c(
    "Control",
    "Tumor"
  )
)

```

---

# 26. Confusion Matrix

The predicted classifications are compared with the actual group labels.

```{r confusion-matrix}

cat("\nConfusion Matrix:\n\n")

confusion_matrix <- table(
  Predicted = pred_class,
  Actual = df$group
)

print(confusion_matrix)

```

---

# 27. Save Risk Scores

The predicted probability for each sample is stored as the `risk_score`.

```{r save-risk-scores}

df$risk_score <- pred

write.csv(
  df,
  file.path(
    result_dir,
    "GSE15932_RiskScores.csv"
  ),
  row.names = FALSE
)

```

---

# 28. Save ROC Curve Coordinates

The sensitivity and specificity values used to construct the ROC curve are saved for further plotting or analysis.

```{r save-roc-coordinates}

roc_df <- data.frame(
  specificity = roc_panel$specificities,
  sensitivity = roc_panel$sensitivities
)

write.csv(
  roc_df,
  file.path(
    result_dir,
    "GSE15932_ROC_curve.csv"
  ),
  row.names = FALSE
)

```

---

# 29. Final Results Summary

```{r final-summary}

cat("\n========================================\n")
cat("GSE15932 EXTERNAL VALIDATION SUMMARY\n")
cat("========================================\n\n")

cat(
  "Number of samples:",
  nrow(df),
  "\n"
)

cat(
  "Control samples:",
  sum(df$group == "Control"),
  "\n"
)

cat(
  "Tumor samples:",
  sum(df$group == "Tumor"),
  "\n"
)

cat(
  "\nGenes included:\n"
)

print(common_genes)

cat(
  "\nMulti-gene AUC:",
  round(
    as.numeric(
      auc(roc_panel)
    ),
    3
  ),
  "\n"
)

cat(
  "\nDeLong 95% CI:\n"
)

print(
  ci.auc(roc_panel)
)

cat(
  "\nOptimal Youden cutoff:",
  round(
    threshold,
    4
  ),
  "\n"
)

cat(
  "\nConfusion Matrix:\n"
)

print(confusion_matrix)

cat(
  "\n========================================\n"
)

cat(
  "External validation completed successfully.\n"
)

cat(
  "========================================\n"
)

```

# 30. Output Files

The following files are generated by this analysis.

### Figures

* `GSE15932_multi_gene_ROC.png`
* `GSE15932_ROC_overlay.png`
* Individual ROC plots for each biomarker in:

`figures/single_gene_ROC/`

### Results

* `GSE15932_single_gene_AUC.csv`
* `GSE15932_RiskScores.csv`
* `GSE15932_ROC_curve.csv`

---

# 31. Interpretation

This analysis evaluates whether the selected biomarker panel can distinguish pancreatic cancer blood samples from healthy controls in the independent **GSE15932** dataset.

The **multi-gene logistic regression model** provides a combined risk score based on the expression levels of the selected biomarkers. Its performance is assessed using the ROC curve and AUC.

Individual ROC analyses allow the diagnostic contribution of each biomarker to be compared with the combined multi-gene panel.

The **Youden index** is used to identify an optimal threshold for converting continuous predicted probabilities into binary classifications of **Tumor** or **Control**.

The confusion matrix provides the resulting classification counts at this threshold.

Importantly, GSE15932 is an **external blood-based validation dataset**, whereas the biomarker discovery analysis was performed using a separate pancreatic cancer dataset. Therefore, the results provide an assessment of whether the selected biomarker signature retains discriminatory performance in an independent blood cohort.
